Hidden GPCR structural transitions addressed by multiple walker supervised molecular dynamics (mwSuMD)

Version 1

Abstract

Abstract

The structural basis for the pharmacology of G protein-coupled receptors (GPCR), the most abundant membrane proteins and the target of about 35% of approved drugs, is still a matter of intense study. What makes GPCRs challenging to study is the inherent flexibility and the metastable nature of interaction with extra- and intracellular partners that drive their effects. Here, we present a molecular dynamics (MD) adaptive sampling algorithm, namely multiple walker supervised molecular dynamics (mwSuMD), to address complex structural transitions involving GPCRs without energy input. By increasing the complexity of the simulated process, we first report the binding and unbinding of the vasopressin peptide from its receptor V2. Successively, we show the stimulatory (Gs) and inhibitory (Gi) G proteins binding to the adrenoreceptor β2 (β2 AR), and the adenosine 1 receptor (A1R), respectively. Then we present the complete transition of the glucagon-like peptide-1 receptor (GLP-1R) from inactive to active, agonist and Gs-bound state, and the GDP release from the activated Gs. Finally, we report the heterodimerization between the adenosine receptor A2 (A2AR) and the dopamine receptor D2 (D2R) and subsequent bivalent ligand binding. We demonstrate that mwSuMD can address, without or with limited energetic bias, complex binding processes such as G protein selectivity and homo- and heterodimerization that are intrinsically linked to the dynamics of the protein and out of reach of classic MD.

Introduction

Supervised molecular dynamics1, 2 (SuMD) is an efficient technique for studying ligand-receptor binding and unbinding pathways; here, we present the multiple walker enhancement (mwSuMD) to study a significantly wider range of structural transitions relevant to the drug design. We validated the method by applying it to G protein-coupled receptors (GPCRs), since their inherent flexibility is essential to their function and because these are the most abundant family of membrane receptors in eukaryotes3 and the target for more than one-third of drugs approved for human use4.

Vertebrate GPCRs are subdivided into five subfamilies (Rhodopsin or class A, Secretin or class B, Glutamate or class C, Adhesion, and Frizzled/Taste2) according to function and sequence5, 6. X-ray and cryo-electron microscopy (cryo-EM) show that GPCRs possess seven transmembrane (TM) helices connected by three extracellular loops (ECLs) and three intracellular loops (ICLs), with an extended and structured N-terminal extracellular domain (ECD) in all subtypes, but class A. The primary function of GPCRs is transducing extracellular chemical signals into the cytosol by binding and activating four G protein families (Gs/olf, Gi/o, G12/13, and Gq/11) responsible for decreasing (Gi/o) or increasing (Gs/olf) cyclic adenosine-3’,5’- monophosphate (cAMP), and generating inositol-1,4,5-triphosphate (IP3) and diacylglycerol (DAG) to increase Ca2+ intracellular levels (Gq)7.

Many aspects of GPCR pharmacology remain elusive: for example, the structural determinants of the selectivity displayed towards specific G proteins or the ability of certain agonists to drive a preferred intracellular signaling pathway over the others (i.e. functional selectivity or bias)8. GPCRs are challenging proteins to characterize experimentally due to their inherent flexibility and the transitory nature of the complexes formed with extracellular and intracellular effectors. Importantly, agonists can allosterically modify the receptor selectivity profile by imprinting unique intracellular conformations from the orthosteric binding site. The mechanism behind these phenomena is one of the outstanding questions in the GPCR field9.

Molecular dynamics (MD) is a powerful computational methodology that predicts the movement and interactions of (bio)molecules in systems of variable complexity, at atomic detail. However, classic MD sampling is limited to the microsecond or, in the best conditions, the millisecond time scale10, 11. For this reason, different algorithms have been designed to speed up the simulation of rare events such as ligand (un)binding and conformational transitions. Amongst the most popular and effective ones, are metadynamics12, accelerated MD (aMD)13, and Gaussian-accelerated MD (GaMD)14, which introduce an energy potential to overcome the energy barriers preventing the complete exploration of the free energy surface, thus de facto bias the simulation. Such methods have been used to study GPCR activation15, 16. Energetically unbiased MD protocols, on the other hand, comprise weighted ensemble MD (weMD)17 and SuMD1, 18, which have largely been applied to (unbinding) small molecules, peptides, and small proteins1, 18–22. Since SuMD is optimized for (un)bindings, we have designed mwSuMD to address more complex conformational transitions and protein-protein associations. GPCRs preferentially couple to very few G proteins out of 23 possible counterparts9, 23. It is increasingly accepted that dynamic and transient interactions determine whether the encounter between a GPCR and a G protein results in productive or unproductive coupling24. MD simulations are considered a useful orthogonal tool for providing working hypotheses and rationalizing existing data on G protein selectivity. However, so far, it has not delivered as expected. Attempts so far have employed energetically biased simulations or have been confined to the Gα subunit15, 16, 23.

We tested mwSuMD on a series of increasingly difficult structural transitions involving both class A and class B1 GPCRs. Firstly, we validated the method on the nonapeptide arginine vasopressin (AVP) by simulating binding (dynamic docking) and unbinding paths from the vasopressin 2 receptor (V2R). Dynamic docking, although more computationally demanding than standard molecular docking, provides insights into the binding mode of ligands in a fully hydrated and flexible environment. Moreover, it informs about binding paths and the complete mechanism of formation leading to an intermolecular complex, delivering in the context of binding kinetics25 and structure-kinetics relationship (SKR) studies26. We then show the Gs and Gi G proteins binding to the adrenoreceptor β2 (β2 AR) and the adenosine 1 receptor (A1R), respectively, starting from different conditions. GPCRs preferentially couple to very few G proteins out of 23 possible counterparts9, 23.

A further case study is the activation of the class B1 GPCR glucagon-like peptide-1 receptor (GLP-1R) by the small molecule PF06882961. GLP-1R is a validated target in type 2 diabetes and probably the best-characterized class B1 GPCR from a structural perspective. GLP-1R is the only class B1 receptor with structurally characterized non-peptidic orthosteric agonists, which makes it a model system for studying the druggability of the entire B1 subfamily. After GLP-1R agonist binding and activation, the coupling of Gs and the release of GDP, which is the rate-limiting step of the G protein activation, was simulated for the first time using an energy-unbiased method.

The last GPCR key process addressed by mwSuMD is the heterodimerization in the membrane between the adenosine receptor A2 (A2AR) and the dopamine receptor D2 (D2R). The A2AR:D2R heterodimer27 is a therapeutic target for neurodegenerative diseases, Parkinson’s disease, and schizophrenia28–30 due to the reciprocal antagonistic allosteric effect between monomers31. A2AR activation reduces the binding affinity of D2R agonists, while A2AR antagonists enhance the dopaminergic tone by decreasing the adenosine negative allosteric modulation on D2R. Heterobivalent ligands able to inhibit A2AR and activate D2R represent a valuable pharmacological tool32 and, in principle, therapeutic options for conditions characterized by a reduction of dopaminergic signaling in the central nervous system. The successive dynamic docking of the heterobivalent ligand compound 2633 to the heterodimer suggested by mwSuMD produced a ternary complex stabilized by lipids and is indicative of the usefulness of mwSuMD for illuminating the molecular events involved in GPCR function.

Results and Discussion

Short mwSuMD time windows improve the AVP dynamic docking prediction

Arginine vasopressin (AVP) is an endogenous hormone (Figure S1a) that mediates antidiuretic effects on the kidney by signaling through three class A GPCR subtypes: V1a and V1b receptors activate phospholipases via Gq/11, while the V2 receptor (V2R) activates adenylyl cyclase by interacting with Gs 34 and is a therapeutic target for hyponatremia, hypertension, and incontinence35. AVP is amphipathic and in the bound state interacts with both polar and hydrophobic V2R residues located on TM helices and ECLs (Figure S1b). Although AVP presents an intramolecular C1-C6 disulfide bond that limits the overall conformational flexibility of the backbone, it has a high number of rotatable bonds, making dynamic docking complicated36. We assessed the performance of mwSuMD and the original version of SuMD in reconstructing the experimental V2R:AVP complex using different settings, simulating a total of 92 binding events (Table S1). As a reference, the AVP RMSD during a classic (unsupervised) equilibrium MD simulation of the X-ray AVP:V2R complex was 3.80 ± 0.52 Å (Figure S2). SuMD1, 18 produced a minimum root mean square deviation (RMSD) to the cryo-EM complex of 4.28 Å, with most of the replicas (i.e., distribution’s mode) having an RMSD close to 10 Å (Figure S3a). mwSuMD, with the same settings (Figure S3b, Table S1) in terms of time window duration (600 ps), metric supervised (the distance between AVP and V2R), and acceptance method (slope) produced slightly more precise results (i.e., distribution’s mode RMSD = 7.90 Å) but similar accuracy (minimum RMSD = 4.60). Supervising the AVP RMSD to the experimental complex rather than the distance (Figure S3c) and using the SMscore (Equation 1) as the acceptance method (Figure S3d) worsened the prediction. Supervising distance and RMSD at the same time (Figure S3e), employing the DMscore (Equation 2), recovered accuracy (minimum RMSD = 4.60 Å) but not precision (distribution mode RMSD = 12.40 Å). Interestingly, decreasing the time window duration from 600 ps to 100 ps impaired the SuMD ability to predict the experimental complex (Figure 1a), but enhanced mwSuMD accuracy and precision (Figure 1b-d). The combination of RMSD as the supervised metric and SMscore produced the best results in terms of minimum RMSD and distribution mode RMSD, 3.85 Å and 4.40 Å, respectively (Figure 1d, Video S1), in agreement with the AVP deviations in the equilibrium MD simulation of the X-ray AVP:V2R complex (Figure S2). These results confirm the inherent complexity of reproducing the AVP:V2R complex via dynamic docking and suggest that short time windows can improve mwSuMD performance on this system. However, it is necessary to know the final bound state to employ the RMSD as the supervised metric, while the distance is required to dynamically dock ligands with unknown bound conformation as previously reported1, 23. Both distance and RMSD-based simulations delivered insights into the binding path and the residues involved along the recognition route. For example, mwSuMD pinpointed V2R residues E184ECL2, P298ECL3, and E303ECL3 (Figure S4a) as involved during AVP binding, although not in contact with the ligand in the orthosteric complex. None of them are yet characterized through mutagenesis studies according to the GPCRdb37.

AVP SuMD and mwSuMD binding simulations to V2R (100 ps time windows).

For each set of settings (a-d) the RMSD of AVP Cα atoms to the cryo-EM structure 7DW9 is reported during the time course of each SuMD (a) or mwSuMD (b-d) replica alongside the RMSD values distribution and the snapshot corresponding to the lowest RMSD values (AVP from the cryo-EM structure 7DW9 is in a cyan stick representation, while AVP from simulations is in a tan stick representation). A complete description of the simulation settings is reported in Table S1 and the Methods section. The dashed red line indicates the AVP RMSD during a classic (unsupervised) equilibrium MD simulation of the X-ray AVP:V2R complex (Figure S2).

Further to binding, a SuMD approach was previously employed to reconstruct the unbinding path of ligands from several GPCRs2, 38. We assessed mwSuMD’s capability to simulate AVP unbinding from V2R. Five mwSuMD and five SuMD replicas were collected using 100 ps time windows (Table S1). Overall, mwSuMD outperformed SuMD in terms of time required to complete a dissociation (Figure S5, Video S2), producing dissociation paths almost 10-fold faster than SuMD. Such rapidity in dissociating inherently produces a limited sampling of metastable states along the pathway, which can be compensated by seeding classic (unsupervised) MD simulations from configurations extracted from the unbinding pathway39, 40. Here, the set of V2R residues involved during the dissociation was comparable to the binding (Figure S4b), though ECL2 and ECL3 were slightly more involved during the association than the dissociation, in analogy with other class A and B GPCRs20, 39.

Binding simulations between G proteins and a class A GPCR

We tested the ability of mwSuMD to simulate the binding between a prototypical class A receptor, the β2 adrenoreceptor (β2 AR), and the stimulatory G protein (Gs), without energy input. mwSuMD simulations started from the intermediate, agonist-bound conformation of the β2-AR and the inactive Gs to resemble pre-coupling conditions. Three mwSuMD replicas were performed by supervising the distance between Gs helix 5 (α5) and β2 AR as well as the RMSD of the intracellular end of TM6 to the fully active state of the receptor (Table S1). To monitor the progression of the simulations, we computed the RMSD of the Cα atoms of the Gα and Gβ subunits to the experimental complex41 (Video S3, Figure 2ab). During two out of three replicas, both Gα and Gβ achieved distance values close to 5 Å (minimum RMSD = 3.94 Å and 3.96 Å respectively), in good agreement with the reference (the β2 AR:Gs complex, PDB 3SN6, Figure 2c). The flexibility of Gsβ is backed by both MD and cryo-EM data suggesting G protein rocking motions around Gsα:receptor interactions20, 42.

G protein binding simulations to β2AR and A1R.

a) RMSD of Gsα to the experimental complex (PDB 3NS6) during three mwSuMD replicas; b) RMSD of Gsβ to the experimental complex (PDB 3NS6) during three mwSuMD replicas; c) superposition of the experimental Gs:β2 AR complex (transparent ribbon) and the MD frame with the lowest Gsα RMSD (3.94 Å); d) RMSD of Giα (residues 243-355) to the A1R experimental complex (PDB 6D9H) during a mwSuMD simulation (red, magnified in the box) and a 1000-ns long classic MD simulation (black); e) two-view superposition of the experimental Gi:A1 R complex (transparent ribbon) and the MD frame with the lowest Giα RMSD (4.82 Å).

Usually, ICL3 of the GPCR and the Gs protein loop hgh4, which is not found in other Gα43, are masked out from deposited cryo-EM structures due to their high flexibility and therefore low resolution. During our simulations, these two loops formed polar intermolecular interactions through R239ICL3, R260ICL3, K235ICL3, and E322hgh4.12, D323hgh4.13. Further transient interactions, not visible in the experiential structures, involved a mix of conserved and unique residues forming hydrogen bonds (Table S2): R63ICL1-E392 α5.24, K2325.71-D378Α5.10, K2355.74-D378 α5.10, K235ICL3-D343H4.13, K2676.29-L394α5, R239ICL3-E314hgh4.04, and S1373.56-D381 α5.13. None of the interactions reported in Table S2 is evident from the experimental β2 AR:Gs complex, implying that mwSuMD can deliver useful working hypotheses for mutagenesis and spectroscopic experiments from out-of-equilibrium simulations. Results also suggest that the Gs binding is driven by a combination of conserved and unique transitory interactions with β2 AR, possibly contributing to G protein selectivity. We speculate that the conserved interactions would be necessary for the binding, regardless of the receptor:G protein couple involved, while the transitory interactions due to the dynamics of G protein binding should produce an effective engagement of the G protein.

A possible pitfall of the above-reported Gs:β2 AR mwSuMD binding simulation is that G proteins bear potential palmitoylation and myristoylation sites that anchor the inactive trimer to the plasma membrane44, 45, de facto restraining possible binding paths to the receptor. Recently, the Gi binding to A1R was simulated by combining the biased methods GaMD with SuMD46 but without taking into account the role played by membrane-anchoring post-translational modifications on the Gi binding pathway. To address this point and test the possible system dependency of mwSuMD, we prepared a different class A GPCR, the adenosine A1 receptor (A1R), and its principal effector, the inhibitory G protein (Gi) considering the Giα residue C3 and Gγ residue C65 as palmitoylated and geranylgeranylated respectively and hence inserted in the membrane. Both classic (unsupervised) and mwSuMD simulations were performed on this system for comparison (Video S4, Figure 2d). In about 50 ns of mwSuMD, the Giα subunit engaged its intracellular binding site on A1R and formed a complex in good agreement with the cryo-EM structure (PDB 6D9H, RMSD ≍ 5 Å). For comparison, 1 μs of cMD did not produce a productive engagement as the Giα remained at RMSD values > 40 Å (Figure 2d), suggesting the effectiveness of mwSuMD in sampling G protein binding rare events without the input of energy. The membrane anchoring affected the overall Gi binding and the final complex, which was rotated compared to the experimental structure due to the lipidation of Giα and Gγ (Figure 2e). This suggests that future, more comprehensive studies of G protein binding and activation should consider several G protein orientations around the receptor as the starting points for mwSuMD simulations, to evaluate as many binding paths as possible

PF06882961 binding, GLP-1R activation, and Gs protein activation

The GLP-1R has been captured by cryo-EM in both the inactive apo (ligand-free) and the active (Gs-bound) conformations and in complex with either peptide or non-peptide agonists47–52. In the inactive apo GLP-1R, residues forming the binding site for the non-peptide agonist PF06882961 are dislocated and scattered due to the structural reorganization of the transmembrane domain (TMD) and extracellular domain (ECD) (Figure S6) that occurs on activation. Moreover, GLP-1R in complex with GLP-1 or different agonists present distinct structural features, even amongst structurally related ligands (Figure S7). This complicates the scenario and suggests divergent recognition mechanisms amongst different agonists. We simulated the binding of PF06882961 using multistep supervision on different metrics of the system (Figure 3) to model the structural hallmark of GLP-1R activation (Video S5, Video S6).

mwSuMD simulation of PF06882961 binding to GLP-1R and receptor activation.

Each panel reports the root mean square deviation (RMSD) to a GLP-1R structural element or the position of the ligand in the active state (top panel), over the time course (all but ECL3 converging to the active state). ECD: extracellular domain; TM: transmembrane helix; ECL: extracellular loop. The mwSuMD simulation was performed with four different settings over 1 microsecond in total. The red dashed lines show the initial RMSD value for reference.

Several metrics were supervised in a consecutive fashion. Firstly, the distance between PF06882961 and the TMD as well as the RMSD of the ECD to the active state (stage 1); secondly, the RMSD of ECD and ECL1 to the active state (stage 2); thirdly, the RMSD of PF06882961 and ECL3 to the active state (stage 3); lastly, only the RMSD of TM6 (residues I345-F367, Cα atoms) to the active state (stage 4). The combination of these supervisions produced a conformational transition of GLP-1R towards the active state (Figure 3, Video S6). Noteworthy, mwSuMD, like any other CV-based technique, requires some degree of knowledge of the system simulated. The sequence of these supervisions was arbitrary and does not necessarily reflect the right order of the steps involved in GLP-1R activation. This kind of planned multistep approach is feasible when the end-point receptor inactive and active structures are available, and the inherent flexibility of different domains is known. In class B GPCRs, the ECD is the most dynamic sub-structure, followed by the ECL1 and ECL3 which display high plasticity during ligand binding20, 53. For this reason, we first supervised these elements of GLP-1R, leaving the bottleneck of activation, TM6 outward movement, as the last step. However, the protocol employed can be tweaked to study how each conformational transition takes place and influences the receptor domains. Structural elements not directly supervised, such as TM1 or TM7, displayed an RMSD reduction to the active state because they were influenced by the movement of supervised helixes or loops. For example, the supervision of ECL3 (stage 3) and TM6 (stage 4) facilitated the spontaneous rearrangement of the ECD to an active-like conformation after the ECD had previously experienced transient high flexibility during stages 2 and 3 (Figure 3).

It is worth noting that the only inactive, almost full-sequence, GLP-1R structure available, was solved with an antibody bound to the ECD. This crystallization strategy for stabilizing the receptor might have forced the ECD in a closed conformation that engages the EC vestibule of GLP-1R and possibly restrained the whole TMD in an altered conformation that deviates from the active conformation. This might explain why the RMSD of the TMD elements monitored during the simulation rarely reach values lower than 3 or 4 Å. During the supervision of ECL3 and PF06882961 (stage 3), we observed a loosening of the intracellular polar interactions that stabilize GLP-1R TM6 in the inactive state. As a result, the subsequent supervision of TM6 (residues I345-F367, Cα atoms) rapidly produced the outward movement of TM6 towards the active state, in the last step of the mwSuMD simulation (stage 4). Taken together, these results suggest a concerted conformational transition for ECD and ECL1 during the binding of PF06882961 and an allosteric effect between ECL3 and the bottom of TM6. While the intracellular polar interactions were destabilized by the ECL3 transition to an active-like conformation (stages 2 and 3), the outward movement of TM6 (stage 4) did not favor the closure of ECL3 towards PF06882961, which appear to be driven by direct interactions between the ligand and R3105.40 or R3807.35. Interestingly, the mwSuMD simulation during Stage 4 (TM6 supervision) sampled a counterclockwise rotation of the helix (Figure 4a), consistent with the GLP-1R cryo-EM structures in the active, Gs-coupled state50, 54. To the best of our knowledge, this is the first time an MD simulation captures the TM6 rotation upon receptor activation as results reported so far are largely limited to the TM6 opening and kinking55.

GLP-1R activation and Gs binding.

a) Rotation of TM6 during GLP-1R activation simulated by mwSuMD. The angle was measured as the dihedral formed by the backbone alpha carbon atoms T353, I357, T362, and I366. The angle values of the inactive and active cryo-EM structures are reported as references; b) RMSD of Gsα to the experimental GLP-1R:Gs complex (PDB 7LCJ) during three mwSuMD replicas; c) PF06882961 MM-GBSA binding energy during Gs binding; d) GDP MM-GBSA binding energy during Gs binding; e-f) sequence of simulated events during the mwSuMD Gs:GLP-1R simulations.

Encouraged by results previously obtained on Gi binding to β2 AR and A1R (Figure 3), we extracted the GLP-1R active conformation described above and simulated the Gs binding to its intracellular side. Starting from the inactive, membrane-anchored Gs, we performed three independent mwSuMD replicas by supervising the distance between Gs helix α5 and GLP-1R residues located at the intracellular binding interface. All three mwSuMD replicas showed the Gs approaching GLP-1R, with two out of three reaching an RMSD of the Gsα subunit close to or less than 10 Å, compared to the experimental complex 7LCI (Figure 4b). Replica 2, in particular, well reproduced the cryo-EM GLP-1R:Gs complex. According to the model of G protein activation, the G protein binding has the effect of allosterically stabilizing the orthosteric agonist in complex with the receptor56 and destabilizing the guanosine-diphosphate (GDP) bound to Gα, triggering its release and exchange with guanosine-triphosphate (GTP)57, upon opening of the G protein alpha-helical domain (AHD). In accordance with this model, PF06882961 and GDP were respectively stabilized and destabilized during the simulated Gs association (Figure 4c,d). The analysis of atomic contacts along the binding path of Gs to GLP-1R highlights a few persistent interactions not observed in the equilibrium MD simulations of GLP-1R:Gs cryo-EM complexes56; in particular, we propose the hidden interaction between D3446.33 and R385 α5 to be pivotal for Gs coupling (Table S3).

We further investigated the Gs activation mechanism by extending Replica 2 (Figure 4b) and supervising the opening of the Gs alpha-helical domain (AHD), which is considered a necessary step to allow GDP release from the Ras-like domain58. We first easily obtained the opening of AHD in the extended Replica 2 and successively supervised the GDP unbinding in further three replicas, seeded after the AHD opening. In one of these three mwSuMD simulations, the nucleotide dissociated from Gs (Figure 4e-h). Video S7 shows the full Gs binding, AHD opening, and GDP release. Interestingly, no conformational changes of the β6-α5 loop happened before or during the GDP dissociation, suggesting that its conformational change as captured in the nucleotide-free GLP-1R:Gs complex (Figure S8) occurs after the GDP release as a result of the loss of binding stabilization, rather than being an initiator of the GDP dissociation. Our simulation is consistent with a major role played by α5 tilting and allosteric communication through internal structural elements like the β2-β3 strands, which weaken both GDP phosphate and purine binding sites as previously suggested by other groups59–61.

The heterodimerization between A2A and D2R, and binding simulations of the heterobivalent ligand compound 26

The current structural model of the A2AR:D2R heterodimer is that TM4 and TM5 from both receptors contribute to the primary dimer interface, although the involvement of TM7 is not ruled out62. Following this interaction model, we first dynamically docked A2AR and D2R in an explicit 1-palmitoyl-2-oleoyl-sn-glycero-3-phosphocholine (POPC) membrane model, then simulated the binding of the heterobivalent compound 2633 (CP26) to the preformed A2AR:D2R heterodimer (Video S8). Since membrane proteins are characterized by slow lateral diffusion63, we encouraged the encounter between A2AR and D2R by inputting energy using metadynamics and adiabatic MD during mwSuMD (hybrid mwSuMD/metadynamics/aMD), followed by 1.5 μs of classic MD (cMD) to relax the system and check the stability of the A2AR:D2R interactions.

During the first 200 ns of simulation with energy bias (Figure 5a,c and Figure S9a), A2AR and D2R rapidly moved close to each other and reached a distance of about 30 Å (computed between centroids), before stabilizing at around 40 Å (Figure 5a). The computed molecular mechanics combined with the Poisson–Boltzmann and surface area continuum solvation (MM-PBSA) binding energy suggested two energy minima (Figure 5c) at these distances. The successive cMD simulation did not produce remarkable changes in the distance between receptors (Figure 5b), although the energy fluctuated before reaching about −10 kcal/mol, at the end of the simulation (Figure 5d). The sharp energy minima after 25 and 150 ns were due to the high number of direct contacts between A2AR and D2R (Figure S9), favored by the energy added to the system. When the input of energy bias was stopped (Figure 5b,d) the POPC residues re-equilibrated at the interface between proteins and mediated intracellular polar interactions between R1504.40 D2R, Y1464.36 D2 and R1995.60 A2A, Y1033.51 A2A as well as extracellular polar interactions between the top of TM4D2, TM5D2 and TM5A2A, TM6A2A (Figure 5f), suggesting that the A2AR:D2R heterodimerization relies on lipids to mediate short-range interactions between receptors.

A2AR:D2R heterodimerization and formation of the ternary complex with C26.

a) Distance between the centroids of A2AR and D2R during the hybrid mwSuMD/metadynamics/aMD/ simulation; b) distance between the centroids of A2AR and D2R during the successive cMD simulation; c) MM-PBSA binding energy between A2AR and D2R during the hybrid metadynamics/aMD/mwSuMD simulation; d) MM-PBSA binding energy between A2AR and D2R during the successive cMD simulation; e) MM-PBSA binding energy between A2AR and D2R during the mwSuMD binding of C26. f) A2AR:D2R heterodimer (white ribbon) after 1.5 μs of cMD; POPC residues (green stick) were involved in polar and hydrophobic interactions; g) extracellular view of the A2AR:D2R:C26 ternary complex (D2R TM2 and TM3 removed for clarity).

The dynamic docking of the heterobivalent ligand C26 further stabilized the A2AR:D2R dimer (Figure 5e), in line with experimental data33. C26 reached the bound state rapidly, inserting the agonist pharmacophore within the D2R orthosteric site (Figure S10, Video S8), while the pyrazole-triazole-pyrimidine scaffold remained in a metastable complex with A2AR, before completely binding the orthosteric site at the end of the simulation (Figure S11, Video S8). In the final state, the long linker between pharmacophores extended over the top of the interface formed by A2AR and D2R at the level of the receptors’ ECL2 (Figure 5g). A network of polar interactions between POPC, Y179A2A, and Y192D2 contributed to stabilizing this ternary complex. Interestingly, the latter residues were pinpointed as important for A2AR:D2R interactions62. From a binding energy perspective, C26 reached the most stable configurations between 80 and 100 ns (Figure S12), before the pyrazole-triazole-pyrimidine component of the ligand completed the binding to A2AR. This suggests some contribution of the linker to the overall stability of the ternary complex with A2AR and D2R. Two out of four mwSuMD replicas produced A2AR:D2R:C26 ternary complexes with C26 engaged both by the orthosteric site of A2AR and D2R, while in the remaining two replicas the A2AR pharmacophore remained stacked on the extracellular vestibule of the receptor, although in the proximity of the binding site (Figure S10). Considering these results, we suggest that the medicinal chemistry optimization of bivalent ligands should involve the tuning of the on-rate kinetic constant kon measured on each monomer of the heterodimer, to differentiate which pharmacophoric scaffold is kinetically more affected by the chemical linking and therefore aiding the surgical improvement of the bivalent ligand’s affinity.

Discussion

Classic MD simulations sample phase space with an efficiency that depends on the energy barrier between neighboring minima. Processes like (un)binding and protein activation require the system to overcome numerous energy barriers, some of which create bottlenecks that slows the transition down to the millisecond, or second, time scale. To overcome some of these limits, we have developed and tested on complex structural events characterizing GPCRs an energetically unbiased adaptive sampling algorithm, namely multiple walker mwSuMD, which is based on traditional SuMD, while drawing on parallel multiple replica methods64, 65,

mwSuMD performed similarly to SuMD for the dynamic docking of AVP to V2R when time windows of 600 ps were employed. Time windows of 100 ps remarkably improved mwSuMD. Usually, dynamic docking is performed to either predict the geometry of complexes or sample the binding path of an already-known intermolecular complex, or both. The RMSD of AVP to the experimental coordinates as the supervised metric produced the best results. Consequently, the RMSD should be the metric of choice to study the binding path of well-known intermolecular complexes. The distance, on the other hand, is necessary when limited structural information about the binding mode is available. In the absence of structural information regarding the final bound state, it is possible to sample numerous binding events employing mwSuMD and evaluate the final bound states rank by applying end-point free energy binding methods like the molecular mechanics energies combined with the Poisson– Boltzmann or generalized Born and surface area continuum solvation (MM/PBSA and MM/GBSA66) models. Our simulations propose that remarkable predictivity can be obtained with distance-driven mwSuMD, as demonstrated by the lowest deviation from the experimental AVP:V2R complex. Remarkably, the dissociation of AVP from V2R was simulated much more rapidly by mwSuMD than by SuMD, suggesting it is an efficient tool for studying the dissociation of ligands from GPCRs.

mwSuMD was further tested on the Gs and Gi binding to β2 AR and A1R, respectively. mwSuMD produced G protein:GPCR complexes in remarkable agreement with experimental structural data without the input of energy in a few hundred nanoseconds when starting from inactive Gs and the intermediate active β2 AR, or a few tens of nanoseconds when considering the active-state A1R and Gi anchored to the plasma membrane through the palmitoylation and the geranylgeranylation of Gαγ44, 45, 67.

We increased the complexity of binding simulations by considering GLP-1R and the non-peptide agonist PF06882961. Using mwSuMD, we obtained a binding of PF06882961in good agreement with the cryo-EM structure, followed by an active-like conformational transition of GLP-1R. The choice of the metrics supervised was driven by structural data available50 and extensive preparatory MD simulations, however, binding routes are possible from either the bulk solvent or the membrane39, 68, 69. Future studies on GLP-1R and other class B1 GPCRs should consider different starting points for the ligand and alternative apo receptor conformations to improve the sampling.

mwSuMD was able to simulate the Gs binding to the active GLP-1R and the subsequent GDP release. To our knowledge, this is the first time the whole sequence of events leading from an inactive GPCR to the GDP release is simulated. Remarkably, this was obtained without biasing the simulations with external energy. Our results suggest that the full rotation and elongation of α5 as in cryo-EM structures could occur after the GDP release, supporting the role of hidden, metastable interactions as the driving force of G protein coupling and selectivity, in accordance with recent work on GLP-1R70.

The final case study was the dimerization process between A2AR and D2R in a membrane model. To speed up the encounter between receptors, we introduced an energy bias in the form of abMD and metadynamics. Although mwSuMD is an unbiased adaptive sampling method, it can be easily coupled to many forms of bias to favor the simulation of energy-requiring processes. Our results suggest a fundamental contribution of the phospholipids on the stabilization of the heterodimer, in agreement with experiments71, 72 but confuting protein-protein molecular docking results frequently predicting extended interfaces between monomers73. mwSuMD was able to dynamically dock the heterobivalent ligand CP26, supporting a stabilizing effect on the A2AR:D2R heterodimer. A complete characterization of the possible interfaces between GPCR monomers, which falls beyond the goal of the present work, should be achieved by preparing different initial unbound states characterized by divergent relative orientations between monomers to dynamically dock.

mwSuMD is suited for a mechanistic description of structural events. For this reason, it well integrates with mutagenesis and kinetics experiments informed by the method, and required for the validation of its results in line with any other computational technique. We note that mwSuMD trajectories, since can describe the sequential states along a transition pathway, can represent a precious backbone for further MD sampling aimed at quantifying or predict the kinetics of the event. Approaches such as path collective variable metadynamics74, Markov State Models75, or machine learning models76, 77 can be informed by mwSuMD. Future work will be addressed in this direction.

In summary, we showcased the applicability domain of mwSuMD to key, but scarcely understood, aspects of GPCR structural biology, pharmacology, and drug design hitherto unaddressed by unbiased simulations. Given the generality and simplicity of its implementation, we anticipate that mwSuMD can be employed to study a wide range of structural phenomena characterizing both membrane and cytosolic proteins. mwSuMD is an ongoing project updated on https://github.com/pipitoludovico/mwSuMD_multiEngine and in the recent implementation can exploit ACEMD78, NAMD79, GROMACS80 and OPENMM81 as MD engine.

Methods

Multiple walker SuMD (mwSuMD) protocol

The supervised MD (SuMD) is an adaptive sampling method82 for speeding up the simulation of binding events between small molecules (or peptides83, 84) and proteins1, 18 without the introduction of any energetic bias. Briefly, during SuMD, a series of short unbiased MD simulations are performed, and after each simulation, the distances between the centers of mass (or the geometrical centers) of the ligand and the predicted binding site (collected at regular time intervals) are fitted to a linear function. If the resulting slope is negative (showing progress towards the target) the next simulation step starts from the last set of coordinates and velocities, otherwise, the simulation is restarted by randomly assigning the atomic velocities.

In the implementation for ACEMD used in this work, mwSuMD needs as input the initial coordinates of the system as a pdb file, the coordinates, and the atomic velocities of the system from the equilibration stage, the topology file of the system, and all the necessary force field parameters. The user can decide to supervise one (X) or two metrics (X’, X’’) of the simulated system over short simulations seeded in batches, called walkers. In the former case, either the slope of the linear function interpolating the metric values or a score can be adopted to decide whether to continue the mwSuMD simulation. When the user decides to supervise two metrics, then a specific score is used. In the present work, distances between centroids, RMSDs, or the number of atomic contacts between two selections were supervised (Table S1). The choice of the metrics is system and problem-dependent, as the RMSD is most useful when the final state is known, while the distance is required when the target state is unknown; details on the scores are given below. The decision to restart or continue mwSuMD after any short simulation is postponed until all the walkers of a batch are collected. The best short simulation is selected and extended by seeding the same number of walkers, with the same duration as the step before.

For each walker, the score for the supervision of a single metric (SMscore) is computed as the square root of the product between the metric value in the last frame (Xlast frame) and the average metric value over the short simulation (X̅):

If the metric is set to decrease (e.g. binding or dimerization) the walker with the lowest SMscore is continued, otherwise (e.g. unbinding or outwards opening of domains), the walker with the highest score is continued. Using the SMscore rather than the slope should give more weight to the final state of each short simulation, as it is the starting point for the successive batch of simulations. Considering the average of the metric should favor short simulations consistently evolving in the desired direction along the metric.

If both X’ and X’’ are set to increase during the mwSuMD simulations, the score for the supervision of two metrics (DMscore) on each walker is computed as follows: Where X’last frame and X’’last frame are the metrics values in the last frame, while X̅ ’batch walkers and X̅ ’’batch walkers represent the average value of the two metrics over all the walkers in the batch. Subtracting the value 1 to the metric ratio ensures that if one of the two metrics from the last frame (X’last frame or X’’last frame) is equal to the average (X̅ ’’batch walkers or X̅ ’’batch walkers) then that metric addend is null and DMscore depends only on the remaining metric. If any of the two metrics is set to decrease, then the corresponding component in Equation 2 is multiplied by −1 to maintain a positive score. Considering the average value of the two metrics over all the walkers rather than only over the considered walker should be more representative of the system evolution along the defined metric. In other words, the information about the metric is taken from all the walkers to better describe the evolution of the system.

The DMScore is designed to preserve some degree of independence between the two metrics supervised. Indeed, if the variation of one of them slows down and gets close to zero, the other metric is still able to drive the system’s evolution. It should be noted that DMScore works at its best if the two metrics have similar variations over time, as in the case of distance and RMSD (both of which are distance-based). Notably, differently from SuMD, when a walker is extended by seeding a new batch of short simulations and the remaining walkers are stopped, the atomic velocities are not reassigned. This allows the simulations to be as short as a few picoseconds if desired, without introducing artifacts due to the thermostat latency to reach the target temperature (usually up to 10-20 ps when a simulation is restarted reassigning the velocities of the atoms).

The current implementation of mwSuMD is for python3 and exploits MDAnalysis85 and MDTRaj86 modules.

Force field, ligands parameters, and general systems preparation

The CHARMM3687, 88/CGenFF 3.0.189–91 force field combination was employed in this work. Initial ligand force field, topology and parameter files were obtained from the ParamChem webserver89. Restrained electrostatic potential (RESP)92 partial charges were assigned to all the non-peptidic small molecules but adrenaline and guanosine-5’-diphosphate (GDP) using Gaussian09 (HF/6-31G* level of theory) and AmberTools20.

Six systems were prepared for MD (Table S1). Hydrogen atoms were added using the pdb2pqr93 and propka94 software (considering a simulated pH of 7.0); the protonation of titratable side chains was checked by visual inspection. The resulting receptors were separately inserted in a 1-palmitoyl-2-oleyl-sn-glycerol-3-phosphocholine (POPC) bilayer (previously built by using the VMD Membrane Builder plugin 1.1, Membrane Plugin, Version 1.1. at: http://www.ks.uiuc.edu/Research/vmd/plugins/membrane/), through an insertion method95. Receptor orientation was obtained by superposing the coordinates on the corresponding structure retrieved from the OPM database96. Lipids overlapping the receptor transmembrane helical bundle were removed and TIP3P water molecules97 were added to the simulation box by means of the VMD Solvate plugin 1.5 (Solvate Plugin, Version 1.5. at <http://www.ks.uiuc.edu/Research/vmd/plugins/solvate/). Finally, overall charge neutrality was reached by adding Na+/Cl- counter ions up to the final concentration of 0.150 M), using the VMD Autoionize plugin 1.3 (Autoionize Plugin, Version 1.3. at <http://www.ks.uiuc.edu/Research/vmd/plugins/autoionize/).

System equilibration and general MD settings

The MD engine ACEMD378 was employed for both the equilibration and productive simulations. The equilibration was achieved in isothermal-isobaric conditions (NPT) using the Berendsen barostat98 (target pressure 1 atm) and the Langevin thermostat99 (target temperature 300 K) with low damping of 1 ps-1. For the equilibration (integration time step of 2 fs): first, clashes between protein and lipid atoms were reduced through 1500 conjugate-gradient minimization steps, then a positional constraint of 1 kcal mol-1 Å-2 on all heavy atoms was gradually released over different time windows: 2 ns for lipid phosphorus atoms, 60 ns for protein atoms other than alpha carbon atoms, 80 ns for alpha carbon atoms; a further 20 ns of equilibration was performed without any positional constraints.

Productive trajectories (Table S1) were computed with an integration time step of 4 fs in the canonical ensemble (NVT). The target temperature was set at 300 K, using a thermostat damping of 0.1 ps-1; the M-SHAKE algorithm100, 101 was employed to constrain the bond lengths involving hydrogen atoms. The cut-off distance for electrostatic interactions was set at 9 Å, with a switching function applied beyond 7.5 Å. Long-range Coulomb interactions were handled using the particle mesh Ewald summation method (PME)102 by setting the mesh spacing to 1.0 Å.

Vasopressin binding simulations

The vasopressin 2 receptor (V2R) in complex with vasopressin (AVP) and the Gs protein103 was retrieved from the Protein Data Bank104 (PDB ID 7DW9). The Gs was removed from the system and the missing residues on ECL2 (G185-G189) were modeled from scratch using Modeller 9.19105. AVP was placed away from V2R in the extracellular bulk and the resulting system was prepared for MD simulations and equilibrated as reported above.

During SuMD simulations, the distance between the centroids of AVP residues C1-Q4 and V2R residues Q96, Q174, Q291, and L312 (Cα atoms only) was supervised over time windows of 600 ps or 100 ps (Table S1). MwSuMD simulations considered the same distance, the RMSD of AVP residues C1-Q4 to the experimental bound complex or the combination of the two during time windows of 600 ps (3 walkers) or 100 ps (10 walkers) (Table S1). Slope, SMscore, or DMscore (see Methods section MwSuMD protocol) was used in the different mwSuMD replicas performed (Table S1). Simulations were stopped after 300 ns (time window duration = 600 ps) or 50 ns (time window duration = 100 ps) of total SuMD or mwSuMD simulation time.

Vasopressin unbinding simulations

The V2R:AVP complex was prepared for MD simulations and equilibrated as reported above. During both SuMD and mwSuMD simulations (Table S1), the distance between the centroids of AVP residues C1-Q4 and V2R residues Q96, Q174, Q291, and L312 (Cα atoms only) was supervised over time windows of 100 ps (10 walkers seeded for mwSuMD simulations). Replicas were stopped when the AVP-V2R distance reached 40 Å.

Gs protein:β2 AR binding simulations

The model of the adrenergic β2 receptor (β2 AR) in an intermediate active state was downloaded from GPCRdb (https://gpcrdb.org/). The full agonist adrenaline (ALE) was inserted in the orthosteric site by superposition with the PDB ID 4LDO (fully active β2 AR)106. The structure of the inactive, GDP-bound Gs protein107 was retrieved from the Protein Data Bank104 (PDB ID 6EG8) and placed in the intracellular bulk. The resulting system (Gs > 50 Å away from (β2 AR) was prepared for MD simulations and equilibrated as reported above. The PDB ID 3SN6 (fully active β2 AR in complex with Gs41) was used as the reference for RMSD computations. Three mwSuMD replicas (Table S1) were performed supervising at the same time the distance between the helix 5 (α5) Gαs residues R385-L395 and the β2 AR residues V31-P330 as well as the RMSD of β2 AR TM6 residues C265-I278 (Cα atoms only) to the fully active state, during 100 ps time windows (5 walkers).

Membrane-anchored Gi protein:A1R simulations

Since the full-length structure of the inactive human Gi protein has not been yet resolved by X-ray or cryo-EM, it was modeled by superimposing the AlphaFold2108 AI models of the Gαi (P63096-F1), Gβ (Q9HAV0-F1), and Gγ (P50151-F1) subunits to the PDB file 6EG8 (a Gs heterotrimer). The resulting homotrimer (without GDP) was processed through Charmm-GUI109 to palmitoylate residue C3Gαi and geranylgeranylate residue C65Gγ 44, 110. The side chains of these two lipidated residues were manually inserted into a 120 x 120 Å POPC membrane and previously built by using the VMD Membrane Builder plugin 1.1, Membrane Plugin, Version 1.1. at: http://www.ks.uiuc.edu/Research/vmd/plugins/membrane/. Lipids overlapping the palmitoyl and geranylgeranyl groups were removed and TIP3P water molecules97 were added to the simulation box by means of the VMD Solvate plugin 1.5 (Solvate Plugin, Version 1.5. at <http://www.ks.uiuc.edu/Research/vmd/plugins/solvate/). Finally, overall charge neutrality was reached by adding Na+/Cl- counter ions up to the final concentration of 0.150 M), using the VMD Autoionize plugin 1.3 (Autoionize Plugin, Version 1.3. at <http://www.ks.uiuc.edu/Research/vmd/plugins/autoionize/). The first stage of equilibration was performed as reported above (Methods section System equilibration and general MD settings) for 120 ns, followed by a second stage in the NVT ensemble for a further 1 μs without any restraints to allow the membrane-anchored heterotrimeric Gi protein to stabilize within the intracellular side of the simulation box. After this two-stage, long equilibration, the active state A1R in complex with adenosine (PDB 6D9H) was manually inserted into the equilibrated membrane above the Gi protein using the corresponding structure retrieved from the OPM database as a reference, and the system further equilibrated for 120 ns as reported above (Methods section System equilibration and general MD settings). The A1R-Gi system was then subjected to both a 1 μs-long classic MD simulation and a mwSuMD simulation (Table S1). During the mwSuMD simulation, the RMSD of helix 5 (α5) Gαs residues 329-354 to the PDB 6D9H was supervised, seeding three walkers of 100 ps each until the productive simulation time reached 50 ns (total simulation time 150 ns).

GLP-1R:PF06882961 binding simulations

The inactive, ligand-free glucagon-like peptide receptor (GLP-1R) was retrieved from the Protein Data Bank104 (PDB ID 6LN2)111. Missing residues in the stalk and ICL2 were modeled with Modeller 9.19. The PF06882961 initial conformation was extracted from the complex with the fully active GLP-1R54 (PDB ID 7LCJ) and placed away from GLP-1R in the extracellular bulk. The resulting system was prepared for MD simulations and equilibrated as reported above. CGenFF dihedral force field parameters of PF06882961 with the highest penalties (dihedrals NG2R51-CG321-CG3C41-CG3C41 (penalty=143.5) and NG2R51-CG321-CG3C41-OG3C51 (penalty=152.4)) were optimized (Figure S13) employing Gaussian09 (geometric optimization and dihedral scan at HF/6-31g(d) level of theory) and the VMD force field toolkit plugin112.

Four classic MD replicas, for a total of 8 μs, were performed on the inactive, ligand-free receptor (prepared for MD simulations and equilibrated as reported above) to assess the possible binding path to the receptor TMD and therefore decide the initial position of PF06882961 in the extracellular bulk of the simulation box. A visual inspection of the trajectories suggested three major conformational changes that could allow ligand access to the TMD (Figure S13). Transitory openings of the ECD (distance Q47ECD - S310ECL2), TM6- TM7 (distance H3636.52 - F3907.45), and TM1-ECL1 (distance E1381.33 and W214ECL1) were observed. Since the opening of TM1-ECL1 was observed in two replicas out of four, we placed the ligand in a favorable position for crossing that region of GLP-1R.

MwSuMD simulations (Table S1) were performed stepwise to dock the ligand within GLP-1R first and then relax the receptor towards the active state. The PF06882961 binding was obtained by supervising at the same time the distance between the ligand and GLP-1R TM7 residues L3797.34-F3817.36, which are part of the orthosteric site (Cα atoms only), and the RMSD of the ECD (residues W33ECD-W120ECD, Cα atoms only) to the active state (PDB ID 7LCJ) until the former distance reached 4 Å. In the second phase of mwSuMD, the RMSD of the ECD (residues W33ECD-W120ECD, Cα atoms only) and the ECL1 to the active state (PDB ID 7LCJ) Cα atoms of residues M2042.74-L2243.27) were supervised until the latter reached less than 4 Å. During the third phase, the RMSD of PF06882961, as well as the RMSD of ECL3 (residues A3686.57-T3787.33, Cα atoms), were supervised until the former reached values lower than 3 Å. In the last mwSuMD step, only the RMSD of TM6 (residues I3456.34-F3676.56, Cα atoms) to the active state (PDB ID 7LCJ) was supervised until less than 5 Å.

Membrane-anchored Gs protein:GLP-1R simulations and GDP dissociation

The PDB 6EG8 was processed through Charmm-GUI109 to palmitoylate residue C3Gαi. The Gβ and Gγ subunits prepared as reported above were superimposed to the PDB file 6EG8 and the original Gβ and Gγ were removed. This step produced a heterotrimeric Gs protein palmitoylated at residue C3Gαs and geranylgeranylated at residue C65Gγ. The resulting system was inserted into a 120 x 120 Å POPC membrane and previously built by using the VMD Membrane Builder plugin 1.1, Membrane Plugin, Version 1.1. at: http://www.ks.uiuc.edu/Research/vmd/plugins/membrane. Lipids overlapping the palmitoyl and geranylgeranyl groups were removed and TIP3P water molecules97 were added to the simulation box by means of the VMD Solvate plugin 1.5 (Solvate Plugin, Version 1.5. at <http://www.ks.uiuc.edu/Research/vmd/plugins/solvate/). Finally, overall charge neutrality was reached by adding Na+/Cl- counter ions up to the final concentration of 0.150 M), using the VMD Autoionize plugin 1.3 (Autoionize Plugin, Version 1.3. at <http://www.ks.uiuc.edu/Research/vmd/plugins/autoionize/). The first stage of equilibration was performed as reported above (Methods section System equilibration and general MD settings) for 120 ns, followed by a second stage in the NVT ensemble for a further 1 μs without any restraints to allow the membrane-anchored heterotrimeric Gs protein to stabilize within the intracellular side of the simulation box. After this two-stage long equilibration, GLP-1R from the final frame of the activation simulation (in complex with PF06882961) was manually inserted into the equilibrated membrane above the Gs protein using the corresponding structure retrieved from the OPM database as a reference, and the system further equilibrated for 120 ns as reported above (Methods section System equilibration and general MD settings). The GLP-1R-Gs system was then subjected to three simulations (Table S1). Each mwSuMD replica was interrupted by 500 ns of classic MD twice, to relax the system during the transition. In the supervised stages, the distance between residues M386-L394 Gαs of helix 5 (α5) and the GLP-1R intracellular residues R1762.46, R3486.37, S3526.41, and N4057.60 was monitored, seeding three walkers of 200 ps each.

The AHD opening was simulated starting from the best GLP-1R:Gs binding mwSuMD replica (Replica 2 in Figure 4) by supervising the distance between AHD residues G70-R199 Gαs and K300-L394Gαs during three walkers of 100 ps each. 300 ns of classic MD was performed to relax the system. Finally, the GDP unbinding was supervised as the distance between GDP and residues E50, K52, T55, K293, and V367 of Gαs; five walkers were used in a 50 ps long mwSuMD simulation.

A2A:D2R heterodimerization

The inactive state A2AR and D2R were retrieved from the Protein Data Bank104 (PDB 5NM4 and 6LUQ, respectively)113, 114. Antagonists bound to the orthosteric site were removed and no modeling of the missing IC loops was attempted. A2AR and D2R were manually placed roughly 40 Å away from each other, on the plane of the membrane, orienting the two receptors to favor the dimerization through the interface formed by TM4 and TM5, as suggested by Borroto-Esquela D. O. et al.62 The resulting system was prepared for MD simulations and equilibrated as reported above.

The heterodimerization between A2AR and D2R was simulated with mwSuMD, seeding batches of three walkers with a duration of 100 ps each (Table S1). During each walker, the distance between TM5 of A2AR and D2R was supervised. At the same time, the distance between the centroids of A2AR and D2R was used as a collective variable for adiabatic MD115 (abMD) and well-tempered metadynamics116, 117 (wtMetaD) performed with Plumed 2.6118. For abMD, a distance target of 30 Å and a force constant of 10000 kJ*mol-1*Å-1) was used, while wtMetaD was performed by seeding Gaussian functions every 1 ps (sigma=1 Å; height=0.837 kJ*mol-1; T=310K) with a bias factor of 30. When the A2AR - D2R distance reached values lower than 40 Å and the first contacts between proteins were formed, the abMD was stopped and wtMetaD continued with a harmonic energy wall at 30 Å to avoid artificial crushing between the receptors due to the added energy bias. When the distance between A2AR and D2R was stable at about 30 Å, the collective variable biased by wtMetaD was set as the number of atomic contacts between A2AR and D2R, until reaching 200 ns of simulation. Finally, to relax the system and challenge the stability of the heterodimer formed during the biased mwSuMD simulation, a 1.5 μs classic MD simulation was performed.

A2AR-D2R heterobitopic ligand binding simulations

The A2AR-D2R heterobivalent ligand compound 2633 was parameterized as reported above and placed in the bulk solvent of the A2AR:D2R complex from the classic MD. Four mwSuMD replicas were collected supervising at the same time the distance between the A2A antagonist pyrazole-triazole-pyrimidine scaffold and the centroid of A2AR residues F168, N253, and A277 (Cα atoms) as well as the distance between the D2 antagonist 4-fluorobenzyl scaffold and the centroids of the Cα of D2R residues C118, F198, and V115 (Cα atoms). Ten walkers of 100 ps were simulated for every mwSuMD batch of replicas.

MD Analysis

Interatomic distances were computed through MDAnalysis85; root mean square deviations (RMSD) were computed using VMD119 and MDAnalysis85.

Interatomic contacts and ligand-protein hydrogen bonds were detected using the GetContacts scripts tool (https://getcontacts.github.io), setting a hydrogen bond donor-acceptor distance of 3.3 Å and an angle value of 120° as geometrical cut-offs. Contacts and hydrogen bond persistency are quantified as the percentage of frames (over all the frames obtained by merging the different replicas) in which protein residues formed contacts or hydrogen bonds with the ligand.

The MMPBSA.py120 script, from the AmberTools20 suite (The Amber Molecular Dynamics Package, at http://ambermd.org/), was used to compute molecular mechanics energies combined with the generalized Born and surface area continuum solvation (MM/GBSA) method or the molecular mechanics Poisson-Boltzmann surface area (MM/PBSA) approach, after transforming the CHARMM psf topology files to an Amber prmtop format using ParmEd (documentation at <http://parmed.github.io/ParmEd/html/index.html). Supplementary Videos were produced employing VMD and avconv (at https://libav.org/avconv.html). Molecular graphics images were produced using the UCSF Chimera121 (v1.14).

Numbering system

Throughout the manuscript, the Ballesteros-Weinstein residues numbering system for class A122 and the Wootten residues numbering system for class B GPCRs123 are adopted.

Supporting information

eLife Assessment

This study provides a methodological report on a modified adaptive sampling approach, multiple walker supervised molecular dynamics (mwSuMD), and its application to G protein-coupled receptors (GPCR), which are the most abundant membrane proteins and key targets for drugs. The mwSuMD approach assists in sampling complex binding processes, leading to some useful findings for GPCR activity, although results may be considered incomplete because the approach may have limited convergence to high-resolution structural data and is lacking other validation. The manuscript explores perhaps too many case studies at the expense of depth of description of methods, reference to existing computational literature, and deeper insight into GPCR activity.

Reviewer #1 (Public Review):

Summary:

The authors investigate ligand and protein-binding processes in GPCRs (including dimerization) by the multiple walker supervised molecular dynamics method. The paper is interesting and it is very well written.

Strengths:

The authors' method is a powerful tool to gain insight into the structural basis for the pharmacology of G protein-coupled receptors.

Weaknesses:

Cholesterol may play a fundamental role in GPCR dimerization (as cited by the authors, Prasanna et al, "Cholesterol-Dependent Conformational Plasticity in GPCR Dimers"). Yet they do not use cholesterol in their simulations of the dimerization.

Reviewer #2 (Public Review):

The study by Deganutti and co-workers is a methodological report on an adaptive sampling approach, multiple walker supervised molecular dynamics (mwSuMD), which represents an improved version of the previous SuMD.

Case-studies concern complex conformational transitions in a number of G protein Coupled Receptors (GPCRs) involving long time-scale motions such as binding-unbinding and collective motions of domains or portions. GPCRs are specialized GEFs (guanine nucleotide exchange factors) of heterotrimeric Gα proteins of the Ras GTPase superfamily. They constitute the largest superfamily of membrane proteins and are of central biomedical relevance as privileged targets of currently marketed drugs.

MwSuMD was exploited to address:

(1) Binding and unbinding of the arginine-vasopressin (AVP) cyclic peptide agonist to the V2 vasopressin receptor (V2R);

(2) Molecular recognition of the β2-adrenergic receptor (β2-AR) and heterotrimeric GDP-bound Gs protein;

(3) Molecular recognition of the A1-adenosine receptor (A1R) and palmitoylated and geranylgeranylated membrane-anchored heterotrimeric GDP-bound Gi protein;

(4) The whole process of GDP release from membrane-anchored heterotrimeric Gs following interaction with the glucagon-like peptide 1 receptor (GLP1R), converted to the active state following interaction with the orthosteric non-peptide agonist danuglipron;

(5) The heterodimerization of D2 dopamine and A2A adenosine receptors (D2R and A2AR, respectively) and binding to a bi-valent ligand.

The mwSuMD method is solid and valuable, has wide applicability, and is compatible with the most world-widely used MD engines. It may be of interest to the computational structural biology community.

The huge amount of high-resolution data on GPCRs makes those systems suitable, although challenging, for method validation and development.

While the approach is less energy-biased than other enhanced sampling methods, knowledge, at the atomic detail, of binding sites/interfaces and conformational states is needed to define the supervised metrics, the higher the resolution of such metrics is the more accurate the outcome is expected to be. The definition of the metrics is a user- and system-dependent process.

The too many and ambitious case-studies undermine the accuracy of the output and reduce the important details needed for a methodological report. In some cases, the available CryoEM structures could have been exploited better.

The most consistent example concerns AVP binding/unbinding to V2R. The consistency with CryoEM data decreases with an increase in the complexity of the simulated process and involved molecular systems (e.g. receptor recognition by membrane-anchored G protein and the process of nucleotide exchange starting from agonist recognition by an inactive-state receptor). The last example, GPCR hetero-dimerization, and binding to a bi-valent ligand, is the most speculative one as it does not rely on high-resolution structural data for metrics supervision.

Reviewer #3 (Public Review):

Summary:

In the present work, Deganutti et al. report a structural study on GPCR functional dynamics using a computational approach called supervised molecular dynamics.

Strengths:

The study has the potential to provide novel insight into GPCR functionality. An example is the interaction between loops of GPCR and G proteins, which are not resolved experimentally, or the interaction between D344 and R385 identified during the Gs coupling by GLP-1R. However, validation of the findings, even computationally through for instance in silico mutagenesis study, is advisable.

Weaknesses:

In its current form, the manuscript seems immature and in particular, the described results grasp only the surface of the complex molecular mechanisms underlying GPCR activation. No significant advance of the existing structural data on GPCR and GPCR/G protein coupling is provided. Most of the results are a reproduction of the previously reported structures.

Author response:

The following is the authors’ response to the original reviews.

Public Reviews:

Reviewer #1 (Public Review):

Summary:

The authors investigate ligand and protein-binding processes in GPCRs (including dimerization) by the multiple walker supervised molecular dynamics method. The paper is interesting and it is very well written.

Strengths:

The authors' method is a powerful tool to gain insight into the structural basis for the pharmacology of G protein-coupled receptors.

Weaknesses:

Cholesterol may play a fundamental role in GPCR dimerization (as cited by the authors, Prasanna et al, "Cholesterol-Dependent Conformational Plasticity in GPCR Dimers"). Yet they do not use cholesterol in their simulations of the dimerization.

We thank Reviewer #1 for the positive comment on mwSuMD.

In the revised version of the manuscript, the section about the A2A/D2 receptors dimerization has been removed because largely speculative. We agree that the lack of cholesterol in those simulations added uncertainty to the presented results.

Reviewer #2 (Public Review):

The study by Deganutti and co-workers is a methodological report on an adaptive sampling approach, multiple walker supervised molecular dynamics (mwSuMD), which represents an improved version of the previous SuMD.

Case-studies concern complex conformational transitions in a number of G protein Coupled Receptors (GPCRs) involving long time-scale motions such as binding-unbinding and collective motions of domains or portions. GPCRs are specialized GEFs (guanine nucleotide exchange factors) of heterotrimeric Gα proteins of the Ras GTPase superfamily. They constitute the largest superfamily of membrane proteins and are of central biomedical relevance as privileged targets of currently marketed drugs.

MwSuMD was exploited to address:

(1) Binding and unbinding of the arginine-vasopressin (AVP) cyclic peptide agonist to the V2 vasopressin receptor (V2R);

(2) Molecular recognition of the β2-adrenergic receptor (β2-AR) and heterotrimeric GDPbound Gs protein;

(3) Molecular recognition of the A1-adenosine receptor (A1R) and palmitoylated and geranylgeranylated membrane-anchored heterotrimeric GDP-bound Gi protein;

(4) The whole process of GDP release from membrane-anchored heterotrimeric Gs following interaction with the glucagon-like peptide 1 receptor (GLP1R), converted to the active state following interaction with the orthosteric non-peptide agonist danuglipron;

(5) The heterodimerization of D2 dopamine and A2A adenosine receptors (D2R and A2AR, respectively) and binding to a bi-valent ligand.

The mwSuMD method is solid and valuable, has wide applicability, and is compatible with the most world-widely used MD engines. It may be of interest to the computational structural biology community.

The huge amount of high-resolution data on GPCRs makes those systems suitable, although challenging, for method validation and development.

While the approach is less energy-biased than other enhanced sampling methods, knowledge, at the atomic detail, of binding sites/interfaces and conformational states is needed to define the supervised metrics, the higher the resolution of such metrics is the more accurate the outcome is expected to be. The definition of the metrics is a user- and system-dependent process.

The too many and ambitious case-studies undermine the accuracy of the output and reduce the important details needed for a methodological report. In some cases, the available CryoEM structures could have been exploited better.

The most consistent example concerns AVP binding/unbinding to V2R. The consistency with CryoEM data decreases with an increase in the complexity of the simulated process and involved molecular systems (e.g. receptor recognition by membrane-anchored G protein and the process of nucleotide exchange starting from agonist recognition by an inactive-state receptor). The last example, GPCR hetero-dimerization, and binding to a bi-valent ligand, is the most speculative one as it does not rely on high-resolution structural data for metrics supervision.

We praise Reviewer #2 for the detailed comment on the manuscript. In this revised version, the hetero-dimerization between A2AR and D2R has been removed. Also, results about GPCR case studies other than GLP-1R have been reduced and downgraded in importance to focus on the fundamental key points of the adaptive sampling method. We agree that the consistency with cryoEM data tends to decrease with an increase in the complexity of the simulated process and involved molecular systems. While it is possible to approximate cryoEM results our unbiased adaptive sampling technique finds its most interesting application in mechanistically unknown out-of-equilibrium processes rather than reproducing known experimental data perfectly. The simulated case studies we present showcase the versatility, speed and consistency of our adaptive method to explore energetically unbiased transitions.

Reviewer #3 (Public Review):

Summary:

In the present work, Deganutti et al. report a structural study on GPCR functional dynamics using a computational approach called supervised molecular dynamics.

Strengths:

The study has the potential to provide novel insight into GPCR functionality. An example is the interaction between loops of GPCR and G proteins, which are not resolved experimentally, or the interaction between D344 and R385 identified during the Gs coupling by GLP-1R. However, validation of the findings, even computationally through for instance in silico mutagenesis study, is advisable.

Weaknesses:

In its current form, the manuscript seems immature and in particular, the described results grasp only the surface of the complex molecular mechanisms underlying GPCR activation. No significant advance of the existing structural data on GPCR and GPCR/G protein coupling is provided. Most of the results are a reproduction of the previously reported structures.

We thank Reviewer #3 for the positive comment on the work. The revised manuscript focuses more on the GLP-1R and Gs case studies. We believe it addresses the weaknesses raised by showing the behaviour of key structural motifs and providing new hypotheses about GDP release.

Reviewer #2 (Recommendations For The Authors):

In this methodological report, Deganutti and co-workers propose an improved version of supervised molecular dynamics (SuMD), named multiple walker SuMD (mwSuMD). Such an adaptive sampling method was challenged in simulations of complex transitions involving GPCRs, which are out of reach by classical MD.

Although less energy-biased than other enhanced sampling methods, mwSuMD requires knowledge of the atomic detail of the ligand-protein or protein-protein binding site/interfaces and the structural hallmarks of the states whose conversion the method is going to address. Such knowledge is, indeed, necessary to define the supervised metrics (e.g. distances, RMSD, etc), which is a user- and system-dependent process.

We classify mwSuMD as an adaptive, rather than enhanced, sampling method as it does not use any energy bias. We agree with the Reviewer that some knowledge of the system is required to productively set up the simulations, but this is the case for almost any MD advanced methods.

The text requires improvement in the essential methodological details and cleaning of those parts is not properly instrumental in method validation.

While attempting to prove the widest possible applicability of the method, the authors exaggerated the number of examples, which, in spite of the increasing complexity were only summarily described. Please, limit the case studies to AVP binding/unbinding to V2R and the whole process of GDP release from membrane-anchored Gs following activation of GLP1R by danuglipron. The latter case, indeed, involves small ligand binding (danuglipron), small ligand dissociation (GDP), receptor activation, and activated receptor binding to membraneanchored G protein and G protein conformational transition instrumental to nucleotide depletion, which is already too much. In this framework, the cases of Gs-β2AR and Gi-A2R recognition are redundant. Most importantly, the case of D2R-A2AR heterodimerization and binding to a bi-valent ligand must be eliminated. The reason is that the case is not entirely based on the mwSuMD and the biased protein-protein interface does not rely on highresolution data (i.e. no structural model of D2R-A2AR dimer has been determined so far). Last but not least, the high intrinsic flexibility of the bi-valent ligand adds further indetermination to the computational experiment. Being too speculative, the case-study does not serve to model validation.

We thank the Reviewer for the suggestion. In the current revised form, the manuscript focuses on AVP binding/unbinding to V2R and the GLP-1R activation, Gs recognition and GDP release.

While eliminating the three case studies mentioned above, the remaining ones should be described more extensively and clearly, highlighting the most productive setup for each system. Incidentally, listing the performance parameters (e.g. distribution mode and minimum RMSD) of each simulation setting in Table S1 is worth doing.

More accuracy in the methodological description is needed.

As for the supervised metrics, the rationale behind the choice of a particular index and whether it is the outcome of a number of trials must be declared and the selected indices must be better defined. Here there are a few examples.

AVP-V2R case. It is not clear why the AVP centroids were computed on residues C1-Q4 (I suppose the Cα-atoms) and not on the Cα-atoms of the whole cyclic part (C1-C6). Along the same line, the choice of the Cα-atoms of four amino acid residues to compute the receptor binding-site centroids requires justification.

We have amended the text to clarify that all the heavy atoms of AVP residues C1-Q4, which are anticipated to bind deep into V2R, were considered alongside V2R residues part of the peptide binding site (Cα atoms only). From our experience, the choice of including side chains or not for the definition of centroids usually does not affect the supervision output. It should only affect the output of mwSuMD simulations based on the RMSD which considers the specific relative distance from the reference. However, a benchmark of the differences produced by divergent selections is beyond the scope of the present work.

GLP1R case. The statement: "Since the opening of TM1-ECL1 was observed in two replicas out of four, we placed the ligand in a favorable position for crossing that region of GLP-1R" is rather weak as a strategy to manually (?) define the input position of the ligand.

As stated in the manuscript, placing the agonist in that position was driven by preliminary 8 μs of classic MD simulations that pointed out the possible path for binding. We agree with the Reviewer that there is still some degree of arbitrarity in it and for this reason, we have not presented structural details of the F06882961 binding path.

As for the supervised metrics, what does it mean "the distance between the ligand and GLP-1R TM7 residues L3797.34-F3817.36"? Was the distance computed between ligand and L379-F381 centroids? Also: "In the supervised stages, the distance between residues M386-L394 Gas of helix 5 (α5) and the GLP-1R intracellular residues R1762.46, R3486.37, S3526.41, and N4057.60 was monitored" was it an inter-centroid distance? Furthermore, "supervising the distance between AHD residues G70-R199 Gas and K300-L394Gas" was it the distance between the centroid of the AHD and the centroid of the C-terminal half of the Ras-like domain? In general, when more than two atoms are involved in distance calculation, please, specify if the distance is inter-centroid.

Also: "During the third phase, the RMSD of PF06882961, as well as the RMSD of ECL3 (residues A3686.57-T3787.33, Ca atoms), were supervised" was the RMSD computed without superimposing the ligand to estimate its roto-translations?

We have added details about the selections used for computing centroids throughout the methods section. For example, all the heavy atoms of F06882961 and the Ca atoms of L379-F381 were considered. RMSD values during GLP-1R activation were computed after superimposition on TM2, ECL1, and TM3 residues 170-240 (Ca atoms). This now has been specified in the text.

The authors considered the 7LCJ GLP1R-danuglipron complex as a fully active reference state instead of considering the receptor from a ternary complex with Gs. The ternary complex (7LCI) was indeed considered as a reference only in simulations of receptor-G protein recognition.

7LCJ and 7LCI are both fully active states. The main difference is that in 7LCJ, Gs coordinates were not deposited. Indeed, their RMSD computed on the TMD Ca atoms and F06882961 is 0.63 Å and 0.54 Å, respectively.

Most importantly, the ternary complex chosen by the authors is not adequate as a reference for simulating the "opening" of the AHD because it bears a miniGs, hence, missing the AHD. In that framework, such an opening is rather vague and was not properly supervised by mwSuMD. The authors must repeat metrics supervisions by using, as a reference, the 6X1A ternary complex, which bears a displaced AHD. This would likely lead to a different path of GDP release.

To the best of our knowledge, there is no evidence that a specific open conformation of the AHD is linked to GDP release. In support, we note that in GPCR ternary complexes, the AHD is usually not modelled because of its high flexibility. The only body of evidence we are aware of is that AHD must open up to allow GDP release. For this reason, we decided to supervise the distance between AHD and the Ras domain without using a reference.

In the statement: "The AHD opening was simulated starting from the best GLP-1R:Gs binding mwSuMD replica" the definition "best binding" requires clarification.

This has been amended, specifying that Replica 2 was considered the “best replica” due to the closed deviation to the cryoEM structure.

As for the case study on β2-AR-Gs recognition, I strongly suggest to eliminate it. However, I'd like to make some comments. The sentence: "the adrenergic β2 receptor (b2 AR) in an intermediate active state was downloaded from GPCRdb (https://gpcrdb.org/)" is vague as it does not indicate what intermediate active state structure was used. Since the goal of the case study was to probe the method in simulating receptor-G protein binding, it would have been better to start with a fully active state of the receptor like the 4LDO structure, employed by the authors only to extract epinephrine.

mwSuMD is designed to provide insights into structural transitions. We started from an intermediate active state of β2-AR in complex with adrenaline because resembling the most populated state stabilised by a full agonist according to NMR studies (DOI:10.1016/j.cell.2015.08.045); the fully-active β2-AR conformation is stabilized only after Gs binding. However, following the Reviewer’s suggestion, we have reduced the presented results for the β2-AR-Gs recognition.

Also in this case, it is not clear if the supervised receptor-G protein distance is between the centroid of the whole 7-helix bundle and the centroid of Gs α5. It is not clear why the TM6 RMSD concerned only the cytosolic end of the helix and did not include the kink region. With that selection, to estimate the outward displacement, RMSD should have been computed without superimposing the considered portion (once all remaining Cα-atoms of the receptors are superimposed).

As the Reviewer pointed out above, some knowledge of the system is required to set up mwSuMD. Using more generic metrics as we did in this case, like the distance between the whole TMD and Gs α5 represents a general approach applicable to other GPCRs, that should allow orthogonal metrics to evolve independently from the supervision.

As now specified in the text, the superimposition for RMSD calculation was performed on residues 40 to 140 Ca atoms, hence not considering TM6.

As for the A1R-Gi recognition, as already stated, I strongly suggest eliminating it. However, I'd like to add some comments. I would discourage the employment of an AlphaFold model for simulations deputed to model validation in general and, in particular, when highresolution structures are available. In this case, the authors would have used the 1GP2 structure of heterotrimeric Gi no matter if from the rat species.

Following the Reviewer’s suggestion, we have dramatically reduced the results presented for the A1R-Gi recognition. We considered 1GP2 for the simulations but H5 lacks the Cterminal six residues and therefore some extent of modelling was still necessary. However, we take the Reviewer’s comment on board and consider it for future work.

Also, the palmitoylation and geranylgeranylation process is quite tortuous and it is not clear why the NVT ensemble was employed in the second stage of equilibration. This is reflected also on the GLP1R case study.

We have amended the text to clarify this passage. The second NVT stage is required for stabilizing the G protein and its orientation in the simulation box. The figure below shows that a plateau of the Ca RMSD during the NVT step was reached after 700 ns for both Gi (black) and Gs (orange).

Here, it is not clear if the RMSD of α5 of Gi was computed with or without superposition.

The RMSD of α5 was computed after superimposing on A1R residues 40-140 Ca atoms (the less flexible region of the receptor). We have now amended the text to report this information.

Reviewer #3 (Recommendations For The Authors):

Points to address:

(1) Root Mean Square Deviation (RMSD) data are often reported as minimum values. It would be useful to provide the average value along the stable part of the trajectories. From the plots in Figure 2ab, it seems that the minimum values reported in the paper are very far from the average ones and thus represent special cases that are seldom reached during simulation. The authors should clarify this point;

For the revised manuscript, we moved Figure 2 to the supplementary material and added average RMSD values for the most notable replicas in Figures 4e and S8a,b. As a reference, in the text, we now report RMSDs from our previous classic MD simulations (https://doi.org/10.1038/s41467-021-27760-0) of Gs:GLP-1R cryoEM structure (Gα = 6.18 ± 2.40 Å; Gβ = 7.22 ± 3.12 Å; Gγ = 9.30 ± 3.65 Å) which show how flexible G proteins bound to GPCRs are and give better context to the RMSD values we measured during mwSuMD simulations.

(2) The RMSD values reported in the paper always refer to single molecules or proteins. It would be useful to also report the RMSD computed over the whole complexes (ligand/GPCR or GPCR/G protein). It would provide a better metric for understanding the general distance between the results and the reference experimental structures;

We have now removed the results sections for A1R and β2 AR to focus on GLP-1R, whose RMSD is analyzed in detail in Figures 2, 3 and 4.

(3) A number of computational works investigated the GPCR/G protein interaction and these studies should be cited and discussed. Examples are the works from Mafi et al. 2023 (doi: 10.1038/s41557-023-01238-6), Fleetwood et al. 2020 (doi: 10.1021/acs.biochem.9b00842), Calderon et al. 2023 and 2024 (doi: 10.1021/acs.jcim.3c00805 and doi: 10.1021/acs.jcim.3c01574), Maria-Solano and Choi 2023 (doi: 10.7554/eLife.90773.1), Mitrovic et al. 2023 (doi: 10.1021/acs.jpcb.3c04897), and D'Amore et al. 2023 (doi: 10.1101/2023.09.14.557711). Many of these works focused on the activation of B2AR and the interaction with its G protein. In addition, Maria-Solano and Choi 2023 and D'Amore et al. 2023 also characterized the rotation of TM6 during the A1R and A2AR activation. Therefore, the claim "To the best of our knowledge, this is the first time an MD simulation captures the TM6 rotation upon receptor activation as results reported so far are largely limited to the TM6 opening and kinking55." is untimely;

We thank the Reviewer for the suggested references. We have added them to the introduction as examples of energy-biased (Calderon et al. 2023 and 2024, Maria-Solano and Choi, Mitrovic et al., D'Amore et al) or adaptive sampling (Fleetwood et al) approaches to GPCR. Since the above articles focus on β2 AR and A1R, we do not discuss them in detail because the results sections for A1R and b2 AR have been drastically reduced in the manuscript.

We note that among the suggested references, only Mafi et al report about a simulated G protein (in a pre-formed complex) and none of the work sampled TM6 rotation without input of energy. However, we have removed the claim from the text.

(4) In the discussion section, the authors claim that a distance-based approach can be employed when the structural data of the endpoints is limited. However, the results obtained from the distance-based protocol during the validation of the approach, which was done using V2R as a reference, are unsatisfying, as acknowledged by the authors themselves. For instance, the RMSD mode value reported for the AVP C alpha atoms with respect to 7DW9 is high, 0.7 nm, whereas the minimum value is 0.38 nm. In addition, some side chains are not oriented in the experimental conformation and might have a different interaction pattern with the receptor if compared with the experimental structure. Considering that in this case the endpoint is known, it is plausible that the performance of the method would degrade even further when data about the target structure is limited. In a real case scenario, the ligand binding mode is unknown and in such a case no RMSD matrix can be used. This represents the major concern of this study that is no prediction is provided, but only - rather inaccurate - reproduction of the known structural data;

The goal of the first part of the work was to compare mwSuMD to SuMD to justify its application on ligand binding using a challenging case study like vasopressin. The general validation of the parent method SuMD as a predictive tool for ligand binding mode has been extensively reported over the years (a few examples: https://doi.org/10.1021/ci400766b ; https://doi.org/10.1021/acs.jcim.5b00702 ; https://doi.org/10.1038/s41598-020-77700-z) and fell beyond the scope of this work.

(5) In the discussion, the authors write "A complete characterization of the possible interfaces between GPCR monomers, which falls beyond the goal of the present work, should be achieved by preparing different initial unbound states characterized by divergent relative orientations between monomers to dynamically dock." It would be useful for the reader to refer to and cite here advanced computational approaches that allow a comprehensive sampling of GPCR dimerization independently from the starting conformation of the receptors. One example is coarse-grained metadynamics as shown in doi: 10.1038/s41467-023-42082-z;

The A2A</sub/D2</sub receptors dimerization has been removed from the manuscript.

(6) In many cases, it is not reported how residues missing from the experimental structures used to model the proteins were reconstructed. This information is important, considering that the authors comment on the results of their calculations on addressing these regions, such as in the case of B2AR. Furthermore, the authors did not report how their initial models were validated. The authors should also explain why they did not model the IC loops of A2AR and D2R;

In the current version of the manuscript, for V2R ECL2 and GLP-1R, we specify that we produced 10 solutions with Modeller and considered the best one in terms of the DOPE score.

The only receptor model used, β2 AR, is now presented as preliminary data focusing on Gs and avoiding any structural detail of the Gs recognition.

As reported above the A2A-D2 dimerization has been removed from the manuscript.

(7) In several cases, the authors state that residues never investigated before play an important role in the interaction between different proteins. An example is provided on page 6 for the B2AR/G protein association. Since this claim is quite significant, it would benefit from validation, at least for further calculations such as in silico mutagenesis studies. Another example is at the end of page 10 where the authors report a hidden interaction between D344 and R385 that is pivotal for Gs coupling by GLP-1R. Is there other evidence supporting this result (previously reported literature data, conservation rate of these residues, etc.)?;

We have removed the supplementary table reporting B2AR/G protein interactions to reduce speculations and added a reference that reports GLP-1 EC50 reduction upon mutation of position 344 to Ala (https://doi.org/10.1021/acscentsci.3c00063).

(8) The authors should provide a deeper discussion about the conformational rearrangement of GPCR and G protein during the coupling. In detail, the conformational changes of microswitich amino acids of GPCR (e.g., PIF, NPxxY, inactivating ionic lock) and alpha helix 5 of G proteins should be discussed in relation to the literature data and experimental structures;

We have removed the A1R and b2 AR results to focus on GLP-1R. Key structural motifs in the polar central network and TM6 kink are analyzed more in detail in Figure 3.

(9) The chronology of the conformational changes of GLP-1R is arbitrarily chosen. During the simulation, the RMSD values reported in Fig. 3 are high and do not demonstrate the full accomplishment of the simulation of the activation process of the receptor;

We agree with the Reviewer that the GLP-1R inactive to active transition was not fully accomplished, compared to other work on class A GPCRs. Unlike class A, class B GPCRs represent a challenging system to work with in silico because inactive starting conformations (e.g 6LN2) are extremely distant from the active one (e.g 7LCJ, 7LCI or 6X18), as demonstrated in Figure S6 for GLP-1R. Here we report the first attempt to model a class B GPCR activation mechanism starting from the inactive state, and even if not fully achieved, we believe it represents state-of-the-art simulations for this class of receptors.

(10) It would be helpful for the reader not familiar with the employed technique that the authors explain in one sentence in the main text the pros and cons of using multiple walkers instead of single walker SuMD;

We thank the Reviewer for the excellent suggestion. In the Discussion, we have now commented that: “more extensive sampling obtainable by seeding multiple parallel short simulations instead of a single simulation for batch”, while in the Methods we explain that “mwSuMD is designed to increase the sampling from a specific configuration by seeding user-decided parallel replicas (walkers) rather than one short simulation as per SuMD. Since one replica for each batch of walkers is always considered productive, mwSuMD gives more control than SuMD on the total wall-clock time used for a simulation. On the flip side, mwSuMD requires multiple GPUs to be the most effective, although any multi-threaded GPU can run more walkers on the same hardware keeping the sampling variety.”.

Minor points to address:

(11) Page 3: the following sentence is duplicated (also found on page 2) "GPCRs preferentially couple to very few G proteins out of 23 possible counterparts";

(12) Page 20: Figure S13 refers to the QM validation of PF06882961 torsional angle, not to the image of the receptor conformational changes, which is instead Figure S14 (please correct figure caption).

We thank the Reviewer for the accurate reading of the manuscript. These typos have been corrected.

Version 2

Abstract

Abstract

The structural basis for the pharmacology of G protein-coupled receptors (GPCRs), the most abundant membrane proteins and the target of about 35% of approved drugs, is still a matter of intense study. What makes GPCRs challenging to study is the inherent flexibility and the metastable nature of interaction with extra- and intracellular partners that drive their effects. Here, we present a molecular dynamics (MD) adaptive sampling algorithm, namely multiple walker supervised molecular dynamics (mwSuMD), to address complex structural transitions involving GPCRs without energy input. We first report the binding and unbinding of the vasopressin peptide from its receptor V2. Successively, we present the complete transition of the glucagon-like peptide-1 receptor (GLP-1R) from inactive to active, agonist and Gs-bound state, and the GDP release from Gs. To our knowledge, this is the first time the whole sequence of events leading from an inactive GPCR to the GDP release is simulated without any energy bias. We demonstrate that mwSuMD can address complex binding processes intrinsically linked to protein dynamics out of reach of classic MD.

Introduction

Supervised molecular dynamics1,2 (SuMD) is an efficient technique for studying ligand-receptor binding and unbinding pathways; here, we present the multiple walker enhancement (mwSuMD) to study a significantly wider range of structural transitions relevant to the drug design. We validated the method by applying it to G protein-coupled receptors (GPCRs), since their inherent flexibility is essential to their function and because these are the most abundant family of membrane receptors in eukaryotes3 and the target for more than one-third of drugs approved for human use4.

Vertebrate GPCRs are subdivided into five subfamilies (Rhodopsin or class A, Secretin or class B, Glutamate or class C, Adhesion, and Frizzled/Taste2) according to function and sequence5,6. X-ray and cryo-electron microscopy (cryo-EM) show that GPCRs possess seven transmembrane (TM) helices connected by three extracellular loops (ECLs) and three intracellular loops (ICLs), with an extended and structured N-terminal extracellular domain (ECD) in all subtypes, but class A. The primary function of GPCRs is transducing extracellular chemical signals into the cytosol by binding and activating four G protein families (Gs/olf, Gi/o, G12/13, and Gq/11) responsible for decreasing (Gi/o) or increasing (Gs/olf) cyclic adenosine-3’,5’-monophosphate (cAMP), and generating inositol-1,4,5-triphosphate (IP3) and diacylglycerol (DAG) to increase Ca2+ intracellular levels (Gq)7.

Many aspects of GPCR pharmacology remain elusive: for example, the structural determinants of the selectivity displayed towards specific G proteins or the ability of certain agonists to drive a preferred intracellular signaling pathway over others (i.e. functional selectivity or bias)8. GPCRs are challenging proteins to characterize experimentally due to their inherent flexibility and the transitory nature of the complexes formed with extracellular and intracellular effectors. Importantly, agonists can allosterically modify the receptor selectivity profile by imprinting unique intracellular conformations from the orthosteric binding site. The mechanism behind these phenomena is one of the outstanding questions in the GPCR field9.

Molecular dynamics (MD) is a powerful computational methodology that predicts the movement and interactions of (bio)molecules in systems of variable complexity, at atomic detail. However, classic MD sampling is limited to the microsecond or, in the best conditions, the millisecond time scale10,11. For this reason, different algorithms have been designed to speed up the simulation of rare events such as ligand (un)binding and conformational transitions. Amongst the most popular and effective ones, are metadynamics12, accelerated MD (aMD)13, and Gaussian-accelerated MD (GaMD)14, which introduce an energy potential to overcome the energy barriers preventing the complete exploration of the free energy surface, thus de facto bias the simulation. Energetically unbiased MD protocols, on the other hand, comprise weighted ensemble MD (weMD)15, swarms approach16, AdaptiveGoal17, and SuMD1,18, which have largely been applied to (unbinding) small molecules, peptides, and small proteins1,18–22. Since SuMD is optimized for (un)bindings, we have designed mwSuMD to address more complex conformational transitions and protein-protein associations. GPCRs preferentially couple to very few G proteins out of 23 possible counterparts9,23. It is increasingly accepted that dynamic and transient interactions determine whether the encounter between a GPCR and a G protein results in productive or unproductive coupling24. MD simulations are considered a useful orthogonal tool for providing working hypotheses and rationalizing existing data on G protein selectivity. However, so far, it has not delivered as expected. Attempts have usually employed energetically biased simulations, have been confined to the Gα subunit, or considered a pre-formed GPCR:G protein complex23,25–27.

Firstly, we validated mwSuMD on the nonapeptide arginine vasopressin (AVP) by simulating binding (dynamic docking) and unbinding paths from the vasopressin 2 receptor (V2R). Dynamic docking, although more computationally demanding than standard molecular docking, provides insights into the binding mode of ligands in a fully hydrated and flexible environment. Moreover, it informs about binding paths and the complete mechanism of formation leading to an intermolecular complex, delivering in the context of binding kinetics28 and structure-kinetics relationship (SKR) studies29. We then studied the class B1 GPCR glucagon-like peptide-1 receptor (GLP-1R) activation by the small molecule PF06882961. GLP-1R is a validated target in type 2 diabetes and probably the best-characterized class B1 GPCR from a structural perspective. GLP-1R is the only class B1 receptor with structurally characterized non-peptidic orthosteric agonists, which makes it a model system for studying the druggability of the entire B1 subfamily. After GLP-1R agonist binding and activation, the coupling of Gs and the release of GDP, the rate-limiting step of the G protein activation, was simulated for the first time using an energy-unbiased method.

These results demonstrate the usefulness of mwSuMD for illuminating the molecular events involved in GPCR function.

Results and discussion

Short mwSuMD time windows improve the AVP dynamic docking prediction

Arginine vasopressin (AVP) is an endogenous hormone (Figure S1a) that mediates antidiuretic effects on the kidney by signaling through three class A GPCR subtypes: V1a and V1b receptors activate phospholipases via Gq/11, while the V2 receptor (V2R) activates adenylyl cyclase by interacting with Gs 30 and is a therapeutic target for hyponatremia, hypertension, and incontinence31. AVP is amphipathic and in the bound state interacts with both polar and hydrophobic V2R residues located on TM helices and ECLs (Figure S1b). Although AVP presents an intramolecular C1-C6 disulfide bond that limits the backbone’s overall conformational flexibility, it has many rotatable bonds, making dynamic docking complicated32. We compared the performance of mwSuMD to the parent algorithm SuMD in reconstructing the experimental V2R:AVP complex using different settings, simulating a total of 92 binding events (Table S1). As a reference, the AVP RMSD during a classic (unsupervised) equilibrium MD simulation of the X-ray AVP:V2R complex was 3.80 ± 0.52 Å (Figure S2). SuMD1,18 produced a minimum root mean square deviation (RMSD) to the cryo-EM complex of 4.28 Å, with most of the replicas (i.e., distribution’s mode) having an RMSD close to 10 Å (Figure S3a). mwSuMD, with the same settings (Figure S3b, Table S1) in terms of time window duration (600 ps), metric supervised (the distance between AVP and V2R), and acceptance method (slope) produced slightly more precise results (i.e., distribution’s mode RMSD = 7.90 Å) but similar accuracy (minimum RMSD = 4.60). Supervising the AVP RMSD to the experimental complex rather than the distance (Figure S3c) and using the SMscore (Equation 1) as the acceptance method (Figure S3d) worsened the prediction. Supervising distance and RMSD at the same time (Figure S3e), employing the DMscore (Equation 2), recovered accuracy (minimum RMSD = 4.60 Å) but not precision (distribution mode RMSD = 12.40 Å). Interestingly, decreasing the time window duration from 600 ps to 100 ps impaired the SuMD ability to predict the experimental complex (Figure 1a), but enhanced mwSuMD accuracy and precision (Figure 1b-d). The combination of RMSD as the supervised metric and SMscore produced the best results in terms of minimum RMSD and distribution mode RMSD, 3.85 Å and 4.40 Å, respectively (Figure 1d, Video S1), in agreement with the AVP deviations in the equilibrium MD simulation of the X-ray AVP:V2R complex (Figure S2). These results confirm the inherent complexity of reproducing the AVP:V2R complex via dynamic docking and suggest that short time windows can improve mwSuMD performance on this system. However, it is necessary to know the final bound state to employ the RMSD as the supervised metric, while the distance is required to dynamically dock ligands with unknown bound conformation as previously reported1,23. Both distance and RMSD-based simulations delivered insights into the binding path and the residues involved along the recognition route. For example, mwSuMD pinpointed V2R residues E184ECL2, P298ECL3, and E303ECL3 (Figure S4a) as involved during AVP binding, although not in contact with the ligand in the orthosteric complex. None of them are yet characterized through mutagenesis studies according to the GPCRdb33.

AVP SuMD and mwSuMD binding simulations to V2R (100 ps time windows).

For each set of settings (a-d) the RMSD of AVP Cα atoms to the cryo-EM structure 7DW9 is reported during the time course of each SuMD (a) or mwSuMD (b-d) replica alongside the RMSD values distribution and the snapshot corresponding to the lowest RMSD values (AVP from the cryo-EM structure 7DW9 is in a cyan stick representation, while AVP from simulations is in a tan stick representation). A complete description of the simulation settings is reported in Table S1 and the Methods section. The dashed red line indicates the AVP RMSD during a classic (unsupervised) equilibrium MD simulation of the X-ray AVP:V2R complex (Figure S2).

Further to binding, a SuMD approach was previously employed to reconstruct the unbinding path of ligands from several GPCRs2,34. We assessed mwSuMD’s capability to simulate AVP unbinding from V2R. Five mwSuMD and five SuMD replicas were collected using 100 ps time windows (Table S1). Overall, mwSuMD outperformed SuMD in terms of time required to complete a dissociation (Figure S5, Video S2), producing dissociation paths almost 10 times faster than SuMD. Such rapidity in dissociating inherently produces a limited sampling of metastable states along the pathway, which can be compensated by seeding classic (unsupervised) MD simulations from configurations extracted from the unbinding pathway35,36. Here, the V2R residues involved during the dissociation were comparable to the binding (Figure S4b), although ECL2 and ECL3 were slightly more involved during the association than the dissociation, in analogy with other class A and B GPCRs20,35.

PF06882961 binding and GLP-1R activation

The GLP-1R has been captured by cryo-electron microscopy (cryo-EM) in both the inactive apo (ligand-free) and the active (Gs-bound) conformations and in complex with either peptide or non-peptide agonists37–42. In the inactive apo GLP-1R, residues forming the binding site for the non-peptide agonist PF06882961 are dislocated and scattered due to the structural reorganization of the transmembrane domain (TMD) and extracellular domain (ECD) (Figure S6) that occurs on activation. Moreover, GLP-1R in complex with GLP-1 or different agonists present distinct structural features, even amongst structurally related ligands (Figure S7). This complicates the scenario and suggests divergent recognition mechanisms amongst different agonists. We simulated the binding of PF06882961 using multistep supervision on different system metrics (Figure 2) to model the structural hallmark of GLP-1R activation (Video S5, Video S6).

mwSuMD simulation of PF06882961 binding to GLP-1R and receptor activation.

Each panel reports the root mean square deviation (RMSD) to the position of the ligand in the active state (top panel) or a GLP-1R structural element, over the time course (all but ECL3 converging to the active state). ECD: extracellular domain; TM: transmembrane helix; ECL: extracellular loop. The mwSuMD simulation was performed with four different settings over 1 microsecond in total. The red dashed lines show the initial RMSD value for reference.

Several metrics were supervised consecutively. Firstly, the distance between PF06882961 and the TMD as well as the RMSD of the ECD to the active state (stage 1); secondly, the RMSD of ECD and ECL1 to the active state (stage 2); thirdly, the RMSD of PF06882961 and ECL3 to the active state (stage 3); lastly, only the RMSD of TM6 (residues I345-F367, Cα atoms) to the active state (stage 4). The combination of these supervisions produced a conformational transition of GLP-1R towards the active state (Figure 2, Video S6). Noteworthy, mwSuMD like any other CV-based technique requires some knowledge of the simulated system. The sequence of these supervisions was arbitrary and does not necessarily reflect the right order of the steps involved in GLP-1R activation. This kind of planned multistep approach is feasible when the end-point receptor inactive and active structures are available, and the inherent flexibility of different domains is known. In class B GPCRs, the ECD is the most dynamic sub-structure, followed by the ECL1 and ECL3 which display high plasticity during ligand binding20,43. For this reason, we first supervised these elements of GLP-1R, leaving the bottleneck of activation, TM6 outward movement, as the last step. However, the protocol employed can be tweaked to study how each conformational transition takes place and influences the receptor domains. Structural elements not directly supervised, such as TM1 or TM7, were influenced by the movement of supervised helixes or loops and therefore displayed an RMSD reduction to the active state. For example, the supervision of ECL3 (stage 3) and TM6 (stage 4) facilitated the spontaneous rearrangement of the ECD to an active-like conformation after the ECD had previously experienced transient high flexibility during stages 2 and 3 (Figure 2). These results suggest a concerted conformational transition for ECD and ECL1 during the binding of PF06882961 and an allosteric effect between ECL3 and the bottom of TM6. While the intracellular polar interactions were destabilized by the ECL3 transition to an active-like conformation (stages 2 and 3), the outward movement of TM6 (stage 4) did not favor the closure of ECL3 towards PF06882961, which appears to be driven by direct interactions between the ligand and R3105.40 or R3807.35. Interestingly, the mwSuMD simulation during Stage 4 (TM6 supervision) sampled a counterclockwise helix rotation (Figure S8a) consistent with the GLP-1R cryo-EM structures in the active, Gs-coupled state40,44.

It is worth noting that 6LN2, the only inactive GLP-1R structure available complete with the ECD, was stabilized and solved with an antibody bound to the ECD. This strategy might have forced the ECD into a closed conformation that engages the EC vestibule of GLP-1R and possibly restrained the whole TMD in an altered conformation that deviates from the physiological conditions. This might explain why the RMSD of the TMD elements monitored during the simulation rarely reach values lower than 3 or 4 Å.

During the supervision of ECL3 and PF06882961 (stage 3), we observed a loosening of the intracellular polar interactions that stabilize GLP-1R TM6 in the inactive state. As a result, the subsequent supervision of TM6 (residues I345-F367, Cα atoms) rapidly produced the outward movement of TM6 towards the active state, in the last step of the mwSuMD simulation (stage 4). A more detailed analysis revealed that the central polar network, which is pivotal for mediating GLP-1 signalling45, and the residues at the TM6 kink level adopted active-like conformations during the final stage of the simulation (Figure 3a). In particular, the central polar network (E3646.53, H3636.52, and Q3947.49) experienced side chain rearrangements (Figure 3b) while extensive TM6 kink dislocation occurred at L3606.49, P3586.48, L3576.47 and I3576.46 (Figure 3c).

GLP-1R key structural motifs during mwSuMD GLP-1R activation.

a) RMSD to the active state GLP-1R (7LCI) of the residues forming the central polar network (magenta) and TM6 kink (orange) during mwSuMD of receptor activation; at the end of the simulations minimum values were reached (dashed square). b) The position of the polar network within the core of TMD (left-hand panel) and comparison between inactive, active and mwSuMD final states for the side chains of the residues forming of the polar network; c) The position of the TM6 kink (right-hand panel) and comparison between inactive, active and mwSuMD final states for the side chains of the residues forming the TM6 kink.

Gs protein binding to GLP-1R and GDP release

We then focused on simulating the Gs binding to GLP1-R, after activation, without energy input. For this purpose, we first tested the binding between the prototypical class A receptor β2 adrenoreceptor (β2 AR), and the stimulatory G protein (Gs) (Video S3, Figure S9a,b) by supervising the distance between Gs helix 5 (α5) and β2AR as well as the RMSD of the intracellular end of TM6 to the fully active state of the receptor (see Supplementary Methods). During two out of three replicas, both Gα and Gβ achieved distance values close to 5 Å (minimum RMSD = 3.94 Å and 3.96 Å respectively), in good agreement with the reference (the β2 AR:Gs complex, PDB 3SN6, Figure S9c). A possible pitfall is that G proteins bear potential palmitoylation and myristoylation sites that anchor the inactive trimer to the plasma membrane46,47, de facto restraining possible binding paths to the receptor. To address this point and test different conditions, we prepared the adenosine A1 receptor (A1R), and its principal effector, the inhibitory G protein (Gi) considering the Giα residue C3 and Gγ residue C65 as palmitoylated and geranylgeranylated respectively and hence inserted in the membrane. Recently, the Gi binding to A1R was simulated by combining the biased methods GaMD with SuMD48 but without considering membrane-anchoring post-translational modifications. Both classic (unsupervised) and mwSuMD simulations were performed on this system for comparison (Video S4, Figure S9d). In about 50 ns of mwSuMD, the Giα subunit engaged its intracellular binding site on A1R and formed a complex in good agreement with the cryo-EM structure (PDB 6D9H, RMSD β 5 Å). For comparison, 1 μs of cMD did not produce a productive engagement as the Giα remained at RMSD values > 40 Å (Figure S9d), suggesting the effectiveness of mwSuMD in sampling G protein binding rare events without the input of energy. The membrane anchoring affected the overall Gi binding and the final complex, which was rotated compared to the experimental structure due to the lipidation of Giα and Gγ (Figure S9e).

Encouraged by results obtained on Gs and Gi binding to β2 AR and A1R (Figure S9), we extracted the GLP-1R active conformation described above and simulated the Gs binding to its intracellular side. Starting from the inactive, membrane-anchored Gs, we performed three independent mwSuMD replicas by supervising the distance between Gs helix α5 and GLP-1R residues located at the intracellular binding interface (Figure 4a,e). All three mwSuMD replicas showed the Gs approaching GLP-1R, with two out of three reaching an RMSD of the Gsα subunit close to or less than 10 Å, compared to the experimental complex 7LCI (Figure 4e). Replica 2, in particular, well reproduced the cryo-EM GLP-1R:Gs complex as suggested by RMSDs to 7LCI of 7.59 ± 1.58 Å, 12.15 ± 2.13 Å, and 13.73 ± 2.24 Å for Gα, Gβ and Gγ respectively. Such values are not far from the RMSDs measured in our previous simulations of GLP-1R in complex with Gs and GLP-149 (Gα = 6.18 ± 2.40 Å; Gβ = 7.22 ± 3.12 Å; Gγ = 9.30 ± 3.65 Å), which indicates overall higher flexibility of Gβ and Gγ compared to Gα, which acts as a sort of fulcrum bound to GLP-1R.

GLP-1R activation and Gs binding.

a-d) sequence of simulated events during the mwSuMD Gs:GLP-1R simulations. e) RMSD of Gsα to the experimental GLP-1R:Gs complex (PDB 7LCJ) during three mwSuMD replicas; the RMSD to the experiential bound conformation (7LCJ) during the second part of Replica2 (red dashed line) is reported for each Gs subunit. RMSDs were computed on Gα residues 11 to 43 and 205 to 394 to the experimental structure 7LCI after superimposition on GLP-1R residues 140-240 Cα atoms. f) RMSD of the αG-α4 loop (purple), the distance between K3426.31 and D323Gα (salmon), and the distance between GDP and D295Gα (green) during Gs binding, AHD opening and GDP dissociation; g) and h) comparison between states extracted from before and after AHD opening. Before AHD opening (a), GLP-1R ICL3 interacted with D323Gα and D295 Gα interacted with GDP; after AHD opening (b) and αG-α4 loop reorganization (curved red arrow), αG and D295 Gα moved away from GDP (straight red arrow), destabilizing its binding to Gs.

According to the model of G protein activation, the G protein binding has the effect of allosterically stabilizing the orthosteric agonist in complex with the receptor49 and destabilizing the guanosine-diphosphate (GDP) bound to Gα, triggering its release and exchange with guanosine-triphosphate (GTP)50, upon opening of the G protein alpha-helical domain (AHD). Following this model, PF06882961 and GDP were respectively stabilized and destabilized during the simulated Gs association (Figure S8b,c). The analysis of atomic contacts along the binding path of Gs to GLP-1R highlights a few persistent interactions not observed in the equilibrium MD simulations of GLP-1R:Gs cryo-EM complexes49; for example, we propose the hidden interaction between D3446.33 and R385α5 to be important for Gs coupling (Table S2), which would explain the GLP-1 EC50 reduction upon mutation of position 344 to Ala51.

Extending Replica 2, we further investigated the Gs activation mechanism by supervising the opening of the Gs alpha-helical domain (AHD), which is considered a necessary step to allow GDP release from the Ras-like domain52. We first easily obtained the opening of AHD (Figure 4b) and successively supervised the GDP unbinding in a further three replicas, seeded after the AHD opening. In one of these three mwSuMD simulations, the nucleotide dissociated from Gs (Figure 4d). Video S7 shows the full Gs binding, AHD opening, and GDP release. mwSuMD suggested several structural changes as implicated in GDP dissociation (Figure 4f-h): i) the AHD opening; ii) the conformational change of αG-α4 loop (residues A303Gα-P332Gα), in concert with a loosening of interactions between D323Gα and K3426.31, and iii) the rupture of the hydrogen bond between GDP and D295Gα triggered by the movement of αG away from the GDP binding site (Figure 4f). The involvement of αG through the αG-α4 during the release of GDP from Gs is supported by hydrogen/deuterium exchange experiments53, while the role of D275Gα has been probed with functional assays on the Gi isoform mutant D272GαiA54. Moreover, a similar αG behavior was recently suggested by MD simulations of the Gs binding pathway to β2AR55. We also note that the αG-α4 loop length and aminoacidic composition diverge among G protein isoforms, further suggesting a role in G protein selectivity consistent with the hypothesis that, in different G proteins, distinct domains of the Gα subunit could be responsible for receptor selectivity56. However, a more subtle allosteric communication through internal structural elements like the β2-β3 strands, prompted by α5 tilting, could have weakened GDP phosphate binding as previously suggested by other groups57–59. Interestingly, no significant conformational changes of the β6-α5 loop happened before or during the GDP dissociation, suggesting that its conformational change as captured in the nucleotide-free GLP-1R:Gs complex (Figure S10) occurs after the GDP release as a result of the loss of binding stabilization, rather than being an initiator of the GDP dissociation.

Discussion

Classic MD simulations sample phase space with an efficiency that depends on the energy barrier between neighboring minima. Processes like (un)binding and protein activation require the system to overcome numerous energy barriers, some creating bottlenecks that slow the transition down to the millisecond, or second, time scale. To reduce some of these limits, we have developed, and tested on complex structural events characterizing GPCRs, an energetically unbiased adaptive sampling algorithm, namely multiple walker SuMD, which is based on traditional SuMD, while drawing on parallel multiple replica methods60,61,

Our simulations propose that remarkable predictivity can be obtained with distance-driven mwSuMD, as demonstrated by the lowest deviation from the experimental AVP:V2R complex. The dissociation of AVP from V2R was simulated much more rapidly by mwSuMD than by SuMD, suggesting it is an efficient tool for studying the dissociation of ligands from GPCRs. This is due to the more extensive sampling obtainable by seeding multiple parallel short simulations instead of a single simulation for batch.

mwSuMD performed similarly to SuMD for the dynamic docking of AVP to V2R when time windows of 600 ps were employed. Time windows of 100 ps remarkably improved mwSuMD. Usually, dynamic docking is performed to either predict the geometry of complexes or sample the binding path of an already-known intermolecular complex, or both. The RMSD of AVP to the experimental coordinates as the supervised metric produced the best results. Consequently, the RMSD should be the metric of choice to study the binding path of well-known intermolecular complexes. The distance, on the other hand, is necessary when limited structural information about the binding mode is available. In the absence of structural information regarding the final bound state, it is possible to sample numerous binding events employing mwSuMD and evaluate the final bound states rank by applying end-point free energy binding methods like the molecular mechanics energies combined with the Poisson– Boltzmann or generalized Born and surface area continuum solvation (MM/PBSA and MM/GBSA62) models.

We increased the complexity of binding simulations by considering GLP-1R and the non-peptide agonist PF06882961. Using mwSuMD, we obtained a binding of PF06882961in good agreement with the cryo-EM structure, followed by an active-like conformational transition of GLP-1R. The choice of the metrics supervised was driven by the structural data available40 and extensive preparatory MD simulations. However, binding routes are possible from either the bulk solvent or the membrane35,63,64. These results show the power of the mwSuMD method, indicating that future applications could include ligand binding from the membrane, or alternative apo receptor conformations to improve the sampling for more difficult receptors.

mwSuMD enabled us to simulate the Gs binding to the active GLP-1R and the subsequent GDP release. Our results suggest a concerted effect on GDP binding produced by AHD opening, αG-α4 loop rearrangement, and αG shift away from the GDP site. The full rotation and elongation of α5 as in cryo-EM structures would occur after the GDP release, supporting the role of hidden, metastable interactions as the driving force of G protein coupling and selectivity, as per recent work on GLP-1R51.

We stress that a complete understanding of a complex molecular event like G protein coupling requires the collection of numerous G protein binding paths and GDP dissociation events. mwSuMD is designed to yield a mechanistic description of structural events. For this reason, it integrates well with mutagenesis and kinetics experiments. We note that mwSuMD trajectories, since they describe the sequential states along a transition pathway, can represent a precious backbone for further MD sampling aimed at quantifying or predicting the kinetics of the transition. Approaches such as path collective variable metadynamics65, Markov State Models66, or machine learning models67,68 can be informed by mwSuMD. We will address these novel opportunities created by mwSuMD in future work.

In summary, we showcased the applicability domain of mwSuMD to key, but scarcely understood, aspects of GPCR structural biology, pharmacology, and drug design hitherto unaddressed by unbiased simulations. Given the generality and simplicity of its implementation, we anticipate that mwSuMD can be employed to study a wide range of structural phenomena characterizing both membrane and cytosolic proteins. mwSuMD is an ongoing project updated on https://github.com/pipitoludovico/mwSuMD_multiEngine and in the recent implementation can exploit ACEMD69, NAMD70, GROMACS71 and OPENMM72 as graphic processing units (GPU)-based MD engines while offering the option to run also on CPUs with GROMACS.

Methods

Multiple walker SuMD (mwSuMD) protocol

The supervised MD (SuMD) is an adaptive sampling method73 for speeding up the simulation of binding events between small molecules (or peptides74,75) and proteins1,18 without the introduction of any energetic bias. Briefly, during SuMD, a series of short unbiased MD simulations are performed, and after each simulation, the distances between the centers of mass (or the geometrical centers) of the ligand and the predicted binding site (collected at regular time intervals) are fitted to a linear function. If the resulting slope is negative (showing progress towards the target) the next simulation step starts from the last set of coordinates and velocities, otherwise, the simulation is restarted by randomly assigning the atomic velocities.

mwSuMD is designed to increase the sampling from a specific configuration by seeding user-decided parallel replicas (walkers) rather than one short simulation as per SuMD. Since one replica for each batch of walkers is always considered productive, mwSuMD gives more control than SuMD on the total wall-clock time used for a simulation. On the flip side, to maximise mwSuMD it is optimal to assign one walker per GPU, requiring multiple GPUs to be effective. However, modern multi-threaded GPUs can still employ mwSuMD with a smaller cost in GPU performance. In the implementation for ACEMD used in this work, mwSuMD needs as input the initial coordinates of the system as a pdb file, the coordinates, and the atomic velocities of the system from the equilibration stage, the topology file of the system, and all the necessary force field parameters. The user can decide to supervise one (X) or two metrics (X’, X’’) of the simulated system over short simulations seeded in batches, called walkers. In the former case, either the slope of the linear function interpolating the metric values or a score can be adopted to decide whether to continue the mwSuMD simulation. When the user decides to supervise two metrics, then a specific score is used. In the present work, distances between centroids, RMSDs, or the number of atomic contacts between two selections were supervised (Table S1). The choice of the metrics is system and problem-dependent, as the RMSD is most useful when the final state is known, while the distance is required when the target state is unknown; details on the scores are given below. The decision to restart or continue mwSuMD after any short simulation is postponed until all the walkers of a batch are collected. The best short simulation is selected and extended by seeding the same number of walkers, with the same duration as the step before.

For each walker, the score for the supervision of a single metric (SMscore) is computed as the square root of the product between the metric value in the last frame (Xlast frame) and the average metric value over the short simulation (X̅): If the metric is set to decrease (e.g. binding or dimerization) the walker with the lowest SMscore is continued, otherwise (e.g. unbinding or outwards opening of domains), the walker with the highest score is continued. Using the SMscore rather than the slope should give more weight to the final state of each short simulation, as it is the starting point for the successive batch of simulations. Considering the average of the metric should favor short simulations consistently evolving in the desired direction along the metric.

If both X’ and X’’ are set to increase during the mwSuMD simulations, the score for the supervision of two metrics (DMscore) on each walker is computed as follows: Where X’last frame and X’’last frame are the metrics values in the last frame, while X̅’batch walkers and X̅’’batch walkers represent the average value of the two metrics over all the walkers in the batch. Subtracting the value 1 to the metric ratio ensures that if one of the two metrics from the last frame (X’last frame or X’’last frame) is equal to the average (X̅’’batch walkers or X̅’’batch walkers) then that metric addend is null and DMscore depends only on the remaining metric. If any of the two metrics is set to decrease, then the corresponding component in Equation 2 is multiplied by -1 to maintain a positive score. Considering the average value of the two metrics over all the walkers rather than only over the considered walker should be more representative of the system evolution along the defined metric. In other words, the information about the metric is taken from all the walkers to better describe the evolution of the system.

The DMScore is designed to preserve some degree of independence between the two metrics supervised. Indeed, if the variation of one of them slows down and gets close to zero, the other metric is still able to drive the system’s evolution. It should be noted that DMScore works at its best if the two metrics have similar variations over time, as in the case of distance and RMSD (both of which are distance-based). Notably, differently from SuMD, when a walker is extended by seeding a new batch of short simulations and the remaining walkers are stopped, the atomic velocities are not reassigned. This allows the simulations to be as short as a few picoseconds if desired, without introducing artifacts due to the thermostat latency to reach the target temperature (usually up to 10-20 ps when a simulation is restarted reassigning the velocities of the atoms).

The current implementation of mwSuMD is for python3 and exploits MDAnalysis76 and MDTRaj77 modules.

Force field, ligands parameters, and general systems preparation

The CHARMM3678,79/CGenFF 3.0.180–82 force field combination was employed in this work. Initial ligand force field, topology and parameter files were obtained from the ParamChem webserver80. Restrained electrostatic potential (RESP)83 partial charges were assigned to all the non-peptidic small molecules but adrenaline and guanosine-5’-diphosphate (GDP) using Gaussian09 (HF/6-31G* level of theory) and AmberTools20.

Six systems were prepared for MD (Table S1). Hydrogen atoms were added using the pdb2pqr84 and propka85 software (considering a simulated pH of 7.0); the protonation of titratable side chains was checked by visual inspection. The resulting receptors were separately inserted in a 1-palmitoyl-2-oleyl-sn-glycerol-3-phosphocholine (POPC) bilayer (previously built by using the VMD Membrane Builder plugin 1.1, Membrane Plugin, Version 1.1. at: http://www.ks.uiuc.edu/Research/vmd/plugins/membrane/), through an insertion method86. Receptor orientation was obtained by superposing the coordinates on the corresponding structure retrieved from the OPM database87. Lipids overlapping the receptor transmembrane helical bundle were removed and TIP3P water molecules88 were added to the simulation box employing the VMD Solvate plugin 1.5 (Solvate Plugin, Version 1.5. at <http://www.ks.uiuc.edu/Research/vmd/plugins/solvate/). Finally, overall charge neutrality was reached by adding Na+/Cl- counter ions up to the final concentration of 0.150 M), using the VMD Autoionize plugin 1.3 (Autoionize Plugin, Version 1.3. at <http://www.ks.uiuc.edu/Research/vmd/plugins/autoionize/).

System equilibration and general MD settings

The MD engine ACEMD369 was employed for both the equilibration and productive simulations. The equilibration was achieved in isothermal-isobaric conditions (NPT) using the Berendsen barostat89 (target pressure 1 atm) and the Langevin thermostat90 (target temperature 300 K) with low damping of 1 ps-1. For the equilibration (integration time step of 2 fs): first, clashes between protein and lipid atoms were reduced through 1500 conjugate-gradient minimization steps, then a positional constraint of 1 kcal mol-1 Å-2 on all heavy atoms was gradually released over different time windows: 2 ns for lipid phosphorus atoms, 60 ns for protein atoms other than alpha carbon atoms, 80 ns for alpha carbon atoms; a further 20 ns of equilibration was performed without any positional constraints.

Productive trajectories (Table S1) were computed with an integration time step of 4 fs in the canonical ensemble (NVT). The target temperature was set at 300 K, using a thermostat damping of 0.1 ps-1; the M-SHAKE algorithm91,92 was employed to constrain the bond lengths involving hydrogen atoms. The cut-off distance for electrostatic interactions was set at 9 Å, with a switching function applied beyond 7.5 Å. Long-range Coulomb interactions were handled using the particle mesh Ewald summation method (PME)93 by setting the mesh spacing to 1.0 Å.

Vasopressin binding simulations

The vasopressin 2 receptor (V2R) in complex with vasopressin (AVP) and the Gs protein94 was retrieved from the Protein Data Bank95 (PDB 7DW9). The Gs was removed from the system and the missing residues on ECL2 (G185-G189) were modeled from scratch using Modeller 9.1996, considering the solution with the lowest DOPE score out of ten conformations produced. AVP was placed away from V2R in the extracellular bulk and the resulting system was prepared for MD simulations and equilibrated as reported above.

During SuMD simulations, the distance between the centroids of AVP residues C1-Q4 (backbone and side chains), anticipated to bind deep into V2R, and the V2R residues lining the peptide binding site Q96, Q174, Q291, and L312 (Cα atoms only) was supervised over time windows of 600 ps or 100 ps (Table S1).mwSuMD simulations considered the same distance, the RMSD of AVP residues C1-Q4 to the experimental bound complex or the combination of the two during time windows of 600 ps (3 walkers) or 100 ps (10 walkers) (Table S1). Slope, SMscore, or DMscore (see Methods section mwSuMD protocol) was used in the different mwSuMD replicas performed (Table S1). Simulations were stopped after 300 ns (time window duration = 600 ps) or 50 ns (time window duration = 100 ps) of total SuMD or mwSuMD simulation time.

Vasopressin unbinding simulations

The V2R:AVP complex was prepared for MD simulations and equilibrated as reported above. During both SuMD and mwSuMD simulations (Table S1), the distance between the centroids of AVP residues C1-Q4 (backbone and side chains) and V2R residues Q96, Q174, Q291, and L312 (Cα atoms only) was supervised over time windows of 100 ps (10 walkers seeded for mwSuMD simulations). Replicas were stopped when the AVP-V2R distance reached 40 Å.

GLP-1R:PF06882961 binding simulations

The inactive, ligand-free glucagon-like peptide receptor (GLP-1R) was retrieved from the Protein Data Bank95 (PDB 6LN2)97. Missing residues in the stalk (129-134) and ICL2 (256 to 263) were modeled with Modeller 9.19, considering the solutions with the lowest DOPE score out of ten conformations produced. The PF06882961 initial conformation was extracted from the complex with the fully active GLP-1R44 (PDB 7LCJ) and placed away from GLP-1R in the extracellular bulk. The resulting system was prepared for MD simulations and equilibrated as reported above. CGenFF dihedral force field parameters of PF06882961 with the highest penalties (dihedrals NG2R51-CG321-CG3C41-CG3C41 (penalty=143.5) and NG2R51-CG321-CG3C41-OG3C51 (penalty=152.4)) were optimized (Figure S11) employing Gaussian09 (geometric optimization and dihedral scan at HF/6-31g(d) level of theory) and the VMD force field toolkit plugin98.

Four classic MD replicas, for a total of 8 μs, were performed on the inactive, ligand-free receptor (prepared for MD simulations and equilibrated as reported above) to assess the possible binding path to the receptor TMD and therefore decide the initial position of PF06882961 in the extracellular bulk of the simulation box. A visual inspection of the trajectories suggested three major conformational changes that could allow ligand access to the TMD (Figure S12). Transitory openings of the ECD (distance Q47ECD - S310ECL2), TM6-TM7 (distance H3636.52 - F3907.45), and TM1-ECL1 (distance E1381.33 and W214ECL1) were observed. Since the opening of TM1-ECL1 was observed in two replicas out of four, we placed the ligand in a favorable position for crossing that region of GLP-1R.

mwSuMD simulations (Table S1) were performed stepwise to dock the ligand within GLP-1R first and then relax the receptor towards the active state. The PF06882961 binding was obtained by supervising at the same time the distance between the ligand’s heavy atoms centroid and the centroid of GLP-1R TM7 residues L3797.34-F3817.36 (Cα atoms only), which are part of the orthosteric site, and the RMSD of the ECD (residues W33ECD-W120ECD, Cα atoms only) to the active state (PDB 7LCJ) until the former distance reached 4 Å. In the second phase of mwSuMD, the RMSD of the ECD (residues W33ECD-W120ECD, Cα atoms only) and the ECL1 to the active state (PDB 7LCJ, Cα atoms of residues M2042.74-L2243.27) were supervised until the latter reached less than 4 Å. During the third phase, the RMSD of PF06882961, as well as the RMSD of ECL3 (residues A3686.57-T3787.33, Cα atoms), were supervised until the former reached values lower than 3 Å. In the last mwSuMD step, only the RMSD of TM6 (residues I3456.34-F3676.56, Cα atoms) to the active state (PDB 7LCJ) was supervised until less than 5 Å. RMSDs were computed after superimposition on TM2, ECL1, and TM3 residues 170-240 (Cα atoms), which is the GLP-1R less flexible part49.

Membrane-anchored Gs protein:GLP-1R simulations and GDP dissociation

The PDB 6EG8 was processed through Charmm-GUI99 to palmitoylate residue C3Gαi and geranylgeranylate residue C65Gγ. The resulting system was inserted into a 120 x 120 Å POPC membrane and previously built by using the VMD Membrane Builder plugin 1.1, Membrane Plugin, Version 1.1. at: http://www.ks.uiuc.edu/Research/vmd/plugins/membrane. Lipids overlapping the palmitoyl and geranylgeranyl groups were removed and TIP3P water molecules88 were added to the simulation box by means of the VMD Solvate plugin 1.5 (Solvate Plugin, Version 1.5. at <http://www.ks.uiuc.edu/Research/vmd/plugins/solvate/). Finally, overall charge neutrality was reached by adding Na+/Cl- counter ions up to the final concentration of 0.150 M), using the VMD Autoionize plugin 1.3 (Autoionize Plugin, Version 1.3. at <http://www.ks.uiuc.edu/Research/vmd/plugins/autoionize/). The first stage of equilibration was performed as reported above (Methods section System equilibration and general MD settings) for 120 ns, followed by a second stage in the NVT ensemble for a further 1 μs without any restraints to allow the membrane-anchored heterotrimeric Gs protein to stabilize within the intracellular side of the simulation box. After this two-stage long equilibration, GLP-1R from the final frame of the activation simulation (in complex with PF06882961) was manually inserted into the equilibrated membrane above the Gs protein using the corresponding structure retrieved from the OPM database as a reference, and the system further equilibrated for 120 ns as reported above (Methods section System equilibration and general MD settings). The GLP-1R-Gs system was then subjected to three simulations (Table S1). Each mwSuMD replica was interrupted by 500 ns of classic MD twice, to relax the system during the transition. In the supervised stages, the distance between residues M386-L394 Gαs (all-atoms centroid) of helix 5 (α5) and the GLP-1R intracellular residues R1762.46, R3486.37, S3526.41, and N4057.60 (Cα atoms only) was monitored, seeding three walkers of 200 ps each.

The AHD opening was simulated starting from the GLP-1R:Gs binding mwSuMD replica with the final lowest Gs RMSD, the lowest PF06882961 binding energy and the highest GDP binding energy (Replica 2 in Figure 4e) by supervising the distance between AHD residues G70-R199 Gαs and K300-L394Gαs (all-atoms centroids) during three walkers of 100 ps each. 300 ns of classic MD was performed to relax the system. Finally, the GDP unbinding was supervised as the distance between GDP (all-atoms centroid) and residues E50Gαs, K52Gαs, T55Gαs, K293Gαs, and V367Gαs (Cα atoms only) of Gαs; five walkers were used in a 50 ps long mwSuMD simulations.

MD Analysis

Interatomic distances were computed through MDAnalysis76; root mean square deviations (RMSD) were computed using VMD100 and MDAnalysis76. Interatomic contacts and ligand-protein hydrogen bonds were detected using the GetContacts scripts tool (https://getcontacts.github.io), setting a hydrogen bond donor-acceptor distance of 3.3 Å and an angle value of 120° as geometrical cut-offs. Contacts and hydrogen bond persistency are quantified as the percentage of frames (over all the frames obtained by merging the different replicas) in which protein residues formed contacts or hydrogen bonds with the ligand.

The MMPBSA.py101 script, from the AmberTools20 suite (The Amber Molecular Dynamics Package, at http://ambermd.org/), was used to compute molecular mechanics energies combined with the generalized Born and surface area continuum solvation (MM/GBSA) method or the molecular mechanics Poisson-Boltzmann surface area (MM/PBSA) approach, after transforming the CHARMM psf topology files to an Amber prmtop format using ParmEd (documentation at <http://parmed.github.io/ParmEd/html/index.html). Supplementary Videos were produced employing VMD and avconv (at https://libav.org/avconv.html). Molecular graphics images were produced using the UCSF Chimera102 (v1.14).

Numbering system

Throughout the manuscript, the Ballesteros-Weinstein residues numbering system for class A103 and the Wootten residues numbering system for class B GPCRs104 are adopted.

eLife Assessment

This study provides a methodological report on a modified adaptive sampling approach, multiple walker supervised molecular dynamics (mwSuMD), and its application to G protein-coupled receptors (GPCRs), which are the most abundant membrane proteins and key targets for drugs. The mwSuMD approach assists in sampling complex binding processes, leading to useful findings for GPCR activity, although results may be considered incomplete, given the high RMSD values reported and lack of validation using experimental data. The manuscript also needs corrected descriptions of high-resolution PDB structures and better relation to existing computational literature.

Reviewer #1 (Public review):

Summary:

The authors investigate ligand and protein-binding processes in GPCRs (including dimerization) by the multiple walker supervised molecular dynamics method. The paper is interesting and it is very well written.

Strengths:

The authors' method is a powerful tool to gain insight on the structural basis for the pharmacology of G protein-coupled receptors.

Reviewer #2 (Public review):

The study by Deganutti and co-workers is a methodological report on an adaptive sampling approach, multiple walker supervised molecular dynamics (mwSuMD), which represents an improved version of the previous SuMD.

Case-studies concern complex conformational transitions in a number of G protein Coupled Receptors (GPCRs) involving long time-scale motions such as binding-unbinding and collective motions of domains or portions. GPCRs are specialized GEFs (guanine nucleotide exchange factors) of heterotrimeric Gα proteins of the Ras GTPase superfamily. They constitute the largest superfamily of membrane proteins and are of central biomedical relevance as privileged targets of currently marketed drugs.

MwSuMD was exploited to address:

a) binding and unbinding of the arginine-vasopressin (AVP) cyclic peptide agonist to the V2 vasopressin receptor (V2R);

b) molecular recognition of the β2-adrenergic receptor (β2-AR) and heterotrimeric GDP-bound Gs protein;

c) molecular recognition of the A1-adenosine receptor (A1R) and palmotoylated and geranylgeranylated membrane-anchored heterotrimeric GDP-bound Gi protein;

d) the whole process of GDP release from membrane-anchored heterotrimeric Gs following interaction with the glucagon-like peptide 1 receptor (GLP1R), converted to the active state following interaction with the orthosteric non-peptide agonist danuglipron.

The revised version has improved clarity and rigor compared to the original also thanks to the reduction in the number of complex case studies treated superficially.

The mwSuMD method is solid and valuable, has wide applicability and is compatible with the most world-widely used MD engines. It may be of interest to the computational structural biology community.

The huge amount of high-resolution data on GPCRs makes those systems suitable, although challenging, for method validation and development.

While the approach is less energy-biased than other enhanced sampling methods, knowledge, at the atomic detail, of binding sites/interfaces and conformational states is needed to define the supervised metrics, the higher the resolution of such metrics is the more accurate the outcome is expected to be. Definition of the metrics is a user- and system-dependent process.

Reviewer #3 (Public review):

Summary:

In the present work Deganutti et al. report a structural study on GPCR functional dynamics using a computational approach called supervised molecular dynamics.

Strengths:

The study has potential to provide novel insight into GPCR functionality. Example is the interaction between D344 and R385 identified during the Gs coupling by GLP-1R. However, validation of the findings, even computationally through for instance in silico mutagenesis study, is advisable.

Weaknesses:

No significant advance of the existing structural data on GPCR and GPCR/G protein coupling is provided. Most of the results are reproductions of the previously reported structures.

Author response:

The following is the authors’ response to the previous reviews

Public Reviews:

Reviewer #1 (Public review):

Summary:

The authors investigate ligand and protein-binding processes in GPCRs (including dimerization) by the multiple walker supervised molecular dynamics method. The paper is interesting and it is very well written.

Strengths:

The authors' method is a powerful tool to gain insight on the structural basis for the pharmacology of G protein-coupled receptors.

We thank the Reviewer for the positive comment on the manuscript and the proposed methods.

Reviewer #2 (Public review):

The study by Deganutti and co-workers is a methodological report on an adaptive sampling approach, multiple walker supervised molecular dynamics (mwSuMD), which represents an improved version of the previous SuMD.

Case-studies concern complex conformational transitions in a number of G protein Coupled Receptors (GPCRs) involving long time-scale motions such as binding-unbinding and collective motions of domains or portions. GPCRs are specialized GEFs (guanine nucleotide exchange factors) of heterotrimeric Gα proteins of the Ras GTPase superfamily. They constitute the largest superfamily of membrane proteins and are of central biomedical relevance as privileged targets of currently marketed drugs.

MwSuMD was exploited to address:

a) binding and unbinding of the arginine-vasopressin (AVP) cyclic peptide agonist to the V2 vasopressin receptor (V2R);

b) molecular recognition of the β2-adrenergic receptor (β2-AR) and heterotrimeric GDPbound Gs protein;

c) molecular recognition of the A1-adenosine receptor (A1R) and palmotoylated and geranylgeranylated membrane-anchored heterotrimeric GDP-bound Gi protein;

d) the whole process of GDP release from membrane-anchored heterotrimeric Gs following interaction with the glucagon-like peptide 1 receptor (GLP1R), converted to the active state following interaction with the orthosteric non-peptide agonist danuglipron.

The revised version has improved clarity and rigor compared to the original also thanks to the reduction in the number of complex case studies treated superficially.

The mwSuMD method is solid and valuable, has wide applicability and is compatible with the most world-widely used MD engines. It may be of interest to the computational structural biology community.

The huge amount of high-resolution data on GPCRs makes those systems suitable, although challenging, for method validation and development.

While the approach is less energy-biased than other enhanced sampling methods, knowledge, at the atomic detail, of binding sites/interfaces and conformational states is needed to define the supervised metrics, the higher the resolution of such metrics is the more accurate the outcome is expected to be. Definition of the metrics is a user- and system-dependent process.

We thank the Reviewer for the positive comment on the revised manuscript and mwSuMD. We agree that the choice of supervised metrics is user- and systemdependent. We aim to improve this aspect in the future with the aid of interpretable machine learning.

Reviewer #3 (Public review):

Summary:

In the present work Deganutti et al. report a structural study on GPCR functional dynamics using a computational approach called supervised molecular dynamics.

Strengths:

The study has potential to provide novel insight into GPCR functionality. Example is the interaction between D344 and R385 identified during the Gs coupling by GLP-1R. However, validation of the findings, even computationally through for instance in silico mutagenesis study, is advisable.

Weaknesses:

No significant advance of the existing structural data on GPCR and GPCR/G protein coupling is provided. Most of the results are reproductions of the previously reported structures.

The method focus of our study (mwSuMD) is an enhancement of the supervised molecular dynamics that allows supervising two metrics at the same time and uses a score, rather than a tabù-like algorithm, for handing the simulation. Further changes are the seeding of parallel short replicas (walkers) rather than a series of short simulations, and the software implementation on different MD engines (e.g. Acemd, OpenMM, NAMD, Gromacs).

We agree with the Reviewer that experimental validation of the findings would be advisable, in line with any computational prediction. We are positive that future studies from our group employing mwSuMD will inform mutagenesis and BRET-based experiments.

Reviewer #2 (Recommendations for the authors):

As for GLP1R, I remain convinced that the 7LCI would have been better as a reference for all simulations than 7LCJ, also because 7LCI holds a slightly more complete ECD.

We agree that 7LCJ would have been a better starting point than 7LCI for simulations because it presents the stalk region, contrary to 7LCJ. However, we do not think it might have influenced the output because the stalk is the most flexible segment of GLP1R, and any initial conformation is usually not retained during MD simulations.

Please, correct everywhere the definition of the 6LN2 structure of GPL1R as a ligand-free or apo, because that structure is indeed bound to a negative allosteric modulator docked on the cytosolic end of helix-6

We thank the reviewer for this precision. The text has been modified accordingly.

As for the beta2-AR, the "full-length" AlphaFold model downloaded from the GPCRdb is not an intermediate active state because it is very similar to the receptor in the 3SN6 complex with Gs. Please, eliminate the inappropriate and speculative adjective "intermediate".

We have changed “intermediate” to “not fully active”, which is less speculative since full activation can be achieved only in the presence of the G protein.

Incidentally, in that model, the C-tail, eliminated by the authors, is completely wrong and occupies the G protein binding site. It is not clear to me the reason why the authors preferred to used an AlphaFold model as an input of simulations rather than a high resolution structural model, e.g. 4LDO. Perhaps, the reason is that all ICL regions, including ICL3, were modeled by AlphaFold even if with low confidence. I disagree with that choice.

We understand the reviewer’s point of view. Should we have simulated an “equilibrium” receptor-ligand complex, we would have made the same choice. However, the conformational changes occurring during a G protein binding are so consistent that the starting conformation of the receptor becomes almost irrelevant as long as a sensate structure is used.

Reviewer #3 (Recommendations for the authors):

The revised version of the manuscript is more concise, focusing only on two systems. However, the authors have responded superficially to the reviewers' comments, merely deleting sections of text, making minor corrections, or adding small additions to the text. In particular, the authors have not addressed the main critical points raised by both Reviewer 2 and Reviewer 3.

For example, the RMSD values for the binding of PF06882961 to GLP-1R remain high, raising doubts about the predictive capabilities of the method, at least for this type of system.

What is the RMSD of the ligand relative to the experimental pose obtained in the simulations? This value must be included in the text.

We have added this piece of information about PF06882961 RMSD in the text, which on page 6 now reads “We simulated the binding of PF06882961, reaching an RMSD to its bound conformation in 7LCJ of 3.79 +- 0.83 Å (computed on the second half of the merged trajectory, superimposing on GLP-1R Ca atoms of TMD residues 150 to 390), using multistep supervision on different system metrics (Figure 2) to model the structural hallmark of GLP-1R activation (Video S5, Video S6).”

Similarly, the activation mechanism of GLP-1R is only partially simulated.

Furthermore, it is not particularly meaningful to justify the high RMSD values of the SuMD simulations for the binding of Gs to GLP-1R by comparing them with those reported under unbiased MD conditions. "Replica 2, in particular, well reproduced the cryo-EM GLP-1R complex as suggested by RMSDs to 7LCI of 7.59{plus minus}1.58Å, 12.15{plus minus}2.13Å, and 13.73{plus minus}2.24Å for Gα, Gβ, and Gγ respectively. Such values are not far from the RMSDs measured in our previous simulations of GLP-1R in complex with Gs and GLP-149 (Gα = 6.18 {plus minus} 2.40 Å; Gβ = 7.22 {plus minus} 3.12 Å; Gγ = 9.30 {plus minus} 3.65 Å), which indicates overall higher flexibility of Gβ and Gγ compared to Gα, which acts as a sort of fulcrum bound to GLP-1R."

Without delving into the accuracy of the various calculations, the authors should acknowledge that comparing protein structures with such high RMSD values has no meaningful significance in terms of convergence toward the same three-dimensional structure.

The text has been edited to accommodate the reviewer’s suggestion and still give the readers the measure of the high flexibility of Gs bound to GLP-1R. It now reads “Such values do not support convergence with the static experimental structure but are not far from the RMSDs measured in our previous simulations of GLP-1R in complex with Gs and GLP-1 (Gα = 6.18 ± 2.40 Å; Gb = 7.22 ± 3.12 Å; Gg = 9.30 ± 3.65 Å), which indicates overall higher flexibility of Gb and Gg compared to Gα, which acts as a sort of fulcrum bound to GLP-1R.”

Have the authors simulated the binding of the Gs protein using the experimentally active structure of GLP-1R in complex with the ligand PF06882961 (PDB ID 7LCJ)? Such a simulation would be useful to assess the quality of the binding simulation of Gs to the GLP1R/PF06882961 complex obtained from the previous SuMD.

We considered performing the Gs binding simulation to the active structure of GLP-1R.

However, the GLP-1R (and other class B receptors) fully active state, as reported in 7LCJ, depends on the presence of the Gs and can be reached only upon effector coupling. Since it is unlikely that the unbound receptor is already in the fully active state, we reasoned that considering it as a starting point for Gs binding simulations would have been an artifact.

An example of the insufficient depth of the authors' replies can be seen in their response: "We note that among the suggested references, only Mafi et al report about a simulated G protein (in a pre-formed complex) and none of the work sampled TM6 rotation without input of energy."

This statement is inaccurate. For instance, D'Amore et al. (Chem 2024, doi: 10.1016/j.chempr.2024.08.004) simulated Gs coupling to A2A as well as TM6 rotation, as did Maria-Solano and Choi (eLife 2023, doi: 10.7554/eLife.90773.1). The former employed path collective variables metadynamics, which is not cited in the introduction or the discussion, despite its relevance to the methodologies mentioned.

Respectfully, our previous reply is correct, as all of the mentioned articles used enhanced (energy-biased) approaches, so the claim “none of the work sampled TM6 rotation without input of energy” stands. The reference to D’Amore et al. (published after the previous round of reviews of this manuscript) has been added to the introduction; we thank the reviewer for pointing it out.

Additionally, SuMD employs a tabu algorithm that applies geometric supervision to the simulation, serving as an alternative approach to enhancing sampling compared to the "input of energy" techniques as called by the authors. A fair discussion should clearly acknowledge this aspect of the SuMD methodology.

We have now specified in the Methods that a tabù-like algorithm is part of SuMD, which, despite being the parent technique of mwSuMD, is not the focus of the present work. We provide extended references for readers interested in SuMD. mwSuMD, on the other hand, does not use a tabù-like algorithm but rather a continuative approach based on a score to select the best walker for each batch, as described in the Methods.

Version 3

Abstract

Abstract

The structural basis for the pharmacology of G protein-coupled receptors (GPCRs), the most abundant membrane proteins and the target of about 35% of approved drugs, is still a matter of intense study. What makes GPCRs challenging to study is the inherent flexibility and the metastable nature of interaction with extra-and intracellular partners that drive their effects. Here, we present a molecular dynamics (MD) adaptive sampling algorithm, namely multiple walker supervised molecular dynamics (mwSuMD), to address complex structural transitions involving GPCRs without energy input. We first report the binding and unbinding of the vasopressin peptide from its receptor V2. Successively, we present the complete transition of the glucagon-like peptide-1 receptor (GLP-1R) from inactive to active, agonist and Gs-bound state, and the GDP release from Gs. To our knowledge, this is the first time the whole sequence of events leading from an inactive GPCR to the GDP release is simulated without any energy bias. We demonstrate that mwSuMD can address complex binding processes intrinsically linked to protein dynamics out of reach of classic MD.

Introduction

Supervised molecular dynamics1,2 (SuMD) is an efficient adaptive sampling technique for studying ligand-receptor binding and unbinding pathways; here, we present the multiple walker enhancement (mwSuMD) to study a significantly wider range of structural transitions relevant to the drug design. We validated the method by applying it to G protein-coupled receptors (GPCRs), since their inherent flexibility is essential to their function and because these are the most abundant family of membrane receptors in eukaryotes3 and the target for more than one-third of drugs approved for human use4.

Vertebrate GPCRs are subdivided into five subfamilies (Rhodopsin or class A, Secretin or class B, Glutamate or class C, Adhesion, and Frizzled/Taste2) according to function and sequence5,6. X-ray and cryo-electron microscopy (cryo-EM) show that GPCRs possess seven transmembrane (TM) helices connected by three extracellular loops (ECLs) and three intracellular loops (ICLs), with an extended and structured N-terminal extracellular domain (ECD) in all subtypes, but class A. The primary function of GPCRs is transducing extracellular chemical signals into the cytosol by binding and activating four G protein families (Gs/olf, Gi/o, G12/13, and Gq/11) responsible for decreasing (Gi/o) or increasing (Gs/olf) cyclic adenosine-3’,5’- monophosphate (cAMP), and generating inositol-1,4,5-triphosphate (IP3) and diacylglycerol (DAG) to increase Ca2+ intracellular levels (Gq)7.

Many aspects of GPCR pharmacology remain elusive: for example, the structural determinants of the selectivity displayed towards specific G proteins or the ability of certain agonists to drive a preferred intracellular signaling pathway over others (i.e. functional selectivity or bias)8. GPCRs are challenging proteins to characterize experimentally due to their inherent flexibility and the transitory nature of the complexes formed with extracellular and intracellular effectors. Importantly, agonists can allosterically modify the receptor selectivity profile by imprinting unique intracellular conformations from the orthosteric binding site. The mechanism behind these phenomena is one of the outstanding questions in the GPCR field9.

Molecular dynamics (MD) is a powerful computational methodology that predicts the movement and interactions of (bio)molecules in systems of variable complexity, at atomic detail. However, classic MD sampling is limited to the microsecond or, in the best conditions, the millisecond time scale10,11. For this reason, different algorithms have been designed to speed up the simulation of rare events such as ligand (un)binding and conformational transitions. Amongst the most popular and effective ones are metadynamics12 and path collective variables metadynamics13, accelerated MD (aMD)14, and Gaussian-accelerated MD (GaMD)15, which introduce an energy potential to overcome the energy barriers preventing the complete exploration of the free energy surface, thus de facto bias the simulation. Energetically unbiased MD protocols, on the other hand, comprise weighted ensemble MD (weMD)16, swarms approach17, AdaptiveGoal18, and SuMD1,19, which have largely been applied to (unbinding) small molecules, peptides, and small proteins1,19–23. Since SuMD is optimized for (un)bindings, we have designed mwSuMD to address more complex conformational transitions and protein-protein associations. GPCRs preferentially couple to very few G proteins out of 23 possible counterparts9,24. It is increasingly accepted that dynamic and transient interactions determine whether the encounter between a GPCR and a G protein results in productive or unproductive coupling25. MD simulations are considered a useful orthogonal tool for providing working hypotheses and rationalizing existing data on G protein selectivity. However, so far, it has not delivered as expected. Attempts have usually employed energetically biased simulations, have been confined to the Gα subunit, or considered a pre-formed GPCR:G protein complex24,26–28.

Firstly, we validated mwSuMD on the nonapeptide arginine vasopressin (AVP) by simulating binding (dynamic docking) and unbinding paths from the vasopressin 2 receptor (V2R). Dynamic docking, although more computationally demanding than standard molecular docking, provides insights into the binding mode of ligands in a fully hydrated and flexible environment. Moreover, it informs about binding paths and the complete mechanism of formation leading to an intermolecular complex, delivering in the context of binding kinetics29 and structure-kinetics relationship (SKR) studies30. We then studied the class B1 GPCR glucagon-like peptide-1 receptor (GLP-1R) activation by the small molecule PF06882961. GLP-1R is a validated target in type 2 diabetes and probably the best-characterized class B1 GPCR from a structural perspective. GLP-1R is the only class B1 receptor with structurally characterized non-peptidic orthosteric agonists, which makes it a model system for studying the druggability of the entire B1 subfamily. After GLP-1R agonist binding and activation, the coupling of Gs and the release of GDP, the rate-limiting step of the G protein activation, was simulated for the first time using an energy-unbiased method.

These results demonstrate the usefulness of mwSuMD for illuminating the molecular events involved in GPCR function.

Results and discussion

Short mwSuMD time windows improve the AVP dynamic docking prediction

Arginine vasopressin (AVP) is an endogenous hormone (Figure S1a) that mediates antidiuretic effects on the kidney by signaling through three class A GPCR subtypes: V1a and V1b receptors activate phospholipases via Gq/11, while the V2 receptor (V2R) activates adenylyl cyclase by interacting with Gs 31 and is a therapeutic target for hyponatremia, hypertension, and incontinence32. AVP is amphipathic and in the bound state interacts with both polar and hydrophobic V2R residues located on TM helices and ECLs (Figure S1b). Although AVP presents an intramolecular C1-C6 disulfide bond that limits the backbone’s overall conformational flexibility, it has many rotatable bonds, making dynamic docking complicated33. We compared the performance of mwSuMD to the parent algorithm SuMD in reconstructing the experimental V2R:AVP complex using different settings, simulating a total of 92 binding events (Table S1). As a reference, the AVP RMSD during a classic (unsupervised) equilibrium MD simulation of the X-ray AVP:V2R complex was 3.80 ± 0.52 Å (Figure S2). SuMD1,19 produced a minimum root mean square deviation (RMSD) to the cryo-EM complex of 4.28 Å, with most of the replicas (i.e., distribution’s mode) having an RMSD close to 10 Å (Figure S3a). mwSuMD, with the same settings (Figure S3b, Table S1) in terms of time window duration (600 ps), metric supervised (the distance between AVP and V2R), and acceptance method (slope) produced slightly more precise results (i.e., distribution’s mode RMSD = 7.90 Å) but similar accuracy (minimum RMSD = 4.60). Supervising the AVP RMSD to the experimental complex rather than the distance (Figure S3c) and using the SMscore (Equation 1) as the acceptance method (Figure S3d) worsened the prediction. Supervising distance and RMSD at the same time (Figure S3e), employing the DMscore (Equation 2), recovered accuracy (minimum RMSD = 4.60 Å) but not precision (distribution mode RMSD = 12.40 Å). Interestingly, decreasing the time window duration from 600 ps to 100 ps impaired the SuMD ability to predict the experimental complex (Figure 1a), but enhanced mwSuMD accuracy and precision (Figure 1b-d). The combination of RMSD as the supervised metric and SMscore produced the best results in terms of minimum RMSD and distribution mode RMSD, 3.85 Å and 4.40 Å, respectively (Figure 1d, Video S1), in agreement with the AVP deviations in the equilibrium MD simulation of the X-ray AVP:V2R complex (Figure S2). These results confirm the inherent complexity of reproducing the AVP:V2R complex via dynamic docking and suggest that short time windows can improve mwSuMD performance on this system. However, it is necessary to know the final bound state to employ the RMSD as the supervised metric, while the distance is required to dynamically dock ligands with unknown bound conformation as previously reported1,24. Both distance and RMSD-based simulations delivered insights into the binding path and the residues involved along the recognition route. For example, mwSuMD pinpointed V2R residues E184ECL2, P298ECL3, and E303ECL3 (Figure S4a) as involved during AVP binding, although not in contact with the ligand in the orthosteric complex. None of them are yet characterized through mutagenesis studies according to the GPCRdb34.

AVP SuMD and mwSuMD binding simulations to V2R (100 ps time windows).

For each set of settings (a-d) the RMSD of AVP Cα atoms to the cryo-EM structure 7DW9 is reported during the time course of each SuMD (a) or mwSuMD (b-d) replica alongside the RMSD values distribution and the snapshot corresponding to the lowest RMSD values (AVP from the cryo-EM structure 7DW9 is in a cyan stick representation, while AVP from simulations is in a tan stick representation). A complete description of the simulation settings is reported in Table S1 and the Methods section. The dashed red line indicates the AVP RMSD during a classic (unsupervised) equilibrium MD simulation of the X-ray AVP:V2R complex (Figure S2).

Further to binding, a SuMD approach was previously employed to reconstruct the unbinding path of ligands from several GPCRs2,35. We assessed mwSuMD’s capability to simulate AVP unbinding from V2R. Five mwSuMD and five SuMD replicas were collected using 100 ps time windows (Table S1). Overall, mwSuMD outperformed SuMD in terms of time required to complete a dissociation (Figure S5, Video S2), producing dissociation paths almost 10 times faster than SuMD. Such rapidity in dissociating inherently produces a limited sampling of metastable states along the pathway, which can be compensated by seeding classic (unsupervised) MD simulations from configurations extracted from the unbinding pathway36,37. Here, the V2R residues involved during the dissociation were comparable to the binding (Figure S4b), although ECL2 and ECL3 were slightly more involved during the association than the dissociation, in analogy with other class A and B GPCRs21,36.

PF06882961 binding and GLP-1R activation

The GLP-1R has been captured by cryo-electron microscopy (cryo-EM) in both the inactive and the active (Gs-bound) conformations and in complex with either peptide or non-peptide agonists38–43. In the inactive GLP-1R, residues forming the binding site for the non-peptide agonist PF06882961 are dislocated and scattered due to the structural reorganization of the transmembrane domain (TMD) and extracellular domain (ECD) (Figure S6) that occurs on activation. Moreover, GLP-1R in complex with GLP-1 or different agonists present distinct structural features, even amongst structurally related ligands (Figure S7). This complicates the scenario and suggests divergent recognition mechanisms amongst different agonists. We simulated the binding of PF06882961, reaching an RMSD to its bound conformation in 7LCJ of 3.79 ± 0.83 Å (computed on the second half of the merged trajectory, superimposing on GLP-1R Cα atoms of TMD residues 150 to 390), using multistep supervision on different system metrics (Figure 2) to model the structural hallmark of GLP-1R activation (Video S5, Video S6).

mwSuMD simulation of PF06882961 binding to GLP-1R and receptor activation.

Each panel reports the root mean square deviation (RMSD) to the position of the ligand in the active state (top panel) or a GLP-1R structural element over the time course (all but ECL3 converging to the active state). ECD: extracellular domain; TM: transmembrane helix; ECL: extracellular loop. The mwSuMD simulation was performed with four different settings over 1 microsecond in total. The red dashed lines show the initial RMSD value for reference.

Several metrics were supervised consecutively. Firstly, the distance between PF06882961 and the TMD as well as the RMSD of the ECD to the active state (stage 1); secondly, the RMSD of ECD and ECL1 to the active state (stage 2); thirdly, the RMSD of PF06882961 and ECL3 to the active state (stage 3); lastly, only the RMSD of TM6 (residues I345-F367, Cα atoms) to the active state (stage 4). The combination of these supervisions produced a conformational transition of GLP-1R towards the active state (Figure 2, Video S6). Noteworthy, mwSuMD, like any other CV-based technique, requires some knowledge of the simulated system. The sequence of these supervisions was arbitrary and does not necessarily reflect the right order of the steps involved in GLP-1R activation. This kind of planned multistep approach is feasible when the end-point receptor inactive and active structures are available, and the inherent flexibility of different domains is known. In class B GPCRs, the ECD is the most dynamic sub-structure, followed by the ECL1 and ECL3, which display high plasticity during ligand binding21,44. For this reason, we first supervised these elements of GLP-1R, leaving the bottleneck of activation, TM6 outward movement, as the last step. However, the protocol employed can be tweaked to study how each conformational transition takes place and influences the receptor domains. Structural elements not directly supervised, such as TM1 or TM7, were influenced by the movement of supervised helixes or loops and therefore displayed an RMSD reduction to the active state. For example, the supervision of ECL3 (stage 3) and TM6 (stage 4) facilitated the spontaneous rearrangement of the ECD to an active-like conformation after the ECD had previously experienced transient high flexibility during stages 2 and 3 (Figure 2). These results suggest a concerted conformational transition for ECD and ECL1 during the binding of PF06882961 and an allosteric effect between ECL3 and the bottom of TM6. While the intracellular polar interactions were destabilized by the ECL3 transition to an active-like conformation (stages 2 and 3), the outward movement of TM6 (stage 4) did not favor the closure of ECL3 towards PF06882961, which appears to be driven by direct interactions between the ligand and R3105.40 or R3807.35. Interestingly, the mwSuMD simulation during Stage 4 (TM6 supervision) sampled a counterclockwise helix rotation (Figure S8a) consistent with the GLP-1R cryo-EM structures in the active, Gs-coupled state41,45.

It is worth noting that 6LN2, the only inactive GLP-1R structure available complete with the ECD, was stabilized and solved with an antibody bound to the ECD. This strategy might have forced the ECD into a closed conformation that engages the EC vestibule of GLP-1R and possibly restrained the whole TMD in an altered conformation that deviates from the physiological conditions. This might explain why the RMSD of the TMD elements monitored during the simulation rarely reach values lower than 3 or 4 Å.

During the supervision of ECL3 and PF06882961 (stage 3), we observed a loosening of the intracellular polar interactions that stabilize GLP-1R TM6 in the inactive state. As a result, the subsequent supervision of TM6 (residues I345-F367, Cα atoms) rapidly produced the outward movement of TM6 towards the active state, in the last step of the mwSuMD simulation (stage 4). A more detailed analysis revealed that the central polar network, which is pivotal for mediating GLP-1 signalling46, and the residues at the TM6 kink level adopted active-like conformations during the final stage of the simulation (Figure 3a). In particular, the central polar network (E3646.53, H3636.52, and Q3947.49) experienced side chain rearrangements (Figure 3b) while extensive TM6 kink dislocation occurred at L3606.49, P3586.48, L3576.47 and I3576.46 (Figure 3c).

GLP-1R key structural motifs during mwSuMD GLP-1R activation.

a) RMSD to the active state GLP-1R (7LCI) of the residues forming the central polar network (magenta) and TM6 kink (orange) during mwSuMD of receptor activation; at the end of the simulations minimum values were reached (dashed square). b) The position of the polar network within the core of TMD (left-hand panel) and comparison between inactive, active and mwSuMD final states for the side chains of the residues forming of the polar network; c) The position of the TM6 kink (right-hand panel) and comparison between inactive, active and mwSuMD final states for the side chains of the residues forming the TM6 kink.

Gs protein binding to GLP-1R and GDP release

We then focused on simulating the Gs binding to GLP1-R, after activation, without energy input. For this purpose, we first tested the binding between the prototypical class A receptor β2 adrenoreceptor (β2 AR), and the stimulatory G protein (Gs) (Video S3, Figure S9a,b) by supervising the distance between Gs helix 5 (α5) and β2AR as well as the RMSD of the intracellular end of TM6 to the fully active state of the receptor (see Supplementary Methods). During two out of three replicas, both Gα and Gβ achieved distance values close to 5 Å (minimum RMSD = 3.94 Å and 3.96 Å respectively), in good agreement with the reference (the β2 AR:Gs complex, PDB 3SN6, Figure S9c). A possible pitfall is that G proteins bear potential palmitoylation and myristoylation sites that anchor the inactive trimer to the plasma membrane47,48, de facto restraining possible binding paths to the receptor. To address this point and test different conditions, we prepared the adenosine A1 receptor (A1R), and its principal effector, the inhibitory G protein (Gi) considering the Giα residue C3 and Gγ residue C65 as palmitoylated and geranylgeranylated respectively and hence inserted in the membrane. Recently, the Gi binding to A1R was simulated by combining the biased methods GaMD with SuMD49 but without considering membrane-anchoring post-translational modifications. Both classic (unsupervised) and mwSuMD simulations were performed on this system for comparison (Video S4, Figure S9d). In about 50 ns of mwSuMD, the Giα subunit engaged its intracellular binding site on A1R and formed a complex in good agreement with the cryo-EM structure (PDB 6D9H, RMSD ≈ 5 Å). For comparison, 1 μs of cMD did not produce a productive engagement as the Giα remained at RMSD values > 40 Å (Figure S9d), suggesting the effectiveness of mwSuMD in sampling G protein binding rare events without the input of energy. The membrane anchoring affected the overall Gi binding and the final complex, which was rotated compared to the experimental structure due to the lipidation of Giα and Gγ (Figure S9e).

Encouraged by results obtained on Gs and Gi binding to β2 AR and A1R (Figure S9), we extracted the GLP-1R active conformation described above and simulated the Gs binding to its intracellular side. Starting from the inactive, membrane-anchored Gs, we performed three independent mwSuMD replicas by supervising the distance between Gs helix α5 and GLP-1R residues located at the intracellular binding interface (Figure 4a,e). All three mwSuMD replicas showed the Gs approaching GLP-1R, with two out of three reaching an RMSD of the Gsα subunit close to or less than 10 Å, compared to the experimental complex 7LCI (Figure 4e). Replica 2, in particular, reproduced the cryo-EM GLP-1R:Gs complex with RMSD values to 7LCI of 7.59 ± 1.58 Å, 12.15 ± 2.13 Å, and 13.73 ± 2.24 Å for Gα, Gβ and Gγ, respectively. Such values do not support convergence with the static experimental structure but are not far from the RMSDs measured in our previous simulations of GLP-1R in complex with Gs and GLP-150 (Gα = 6.18 ± 2.40 Å; Gβ = 7.22 ± 3.12 Å; Gγ = 9.30 ± 3.65 Å), which indicates overall higher flexibility of Gβ and Gγ compared to Gα, which acts as a sort of fulcrum bound to GLP-1R.

GLP-1R activation and Gs binding.

a-d) sequence of simulated events during the mwSuMD Gs:GLP-1R simulations. e) RMSD of Gsα to the experimental GLP-1R:Gs complex (PDB 7LCJ) during three mwSuMD replicas; the RMSD to the experiential bound conformation (7LCJ) during the second part of Replica2 (red dashed line) is reported for each Gs subunit. RMSDs were computed on Gα residues 11 to 43 and 205 to 394 to the experimental structure 7LCI after superimposition on GLP-1R residues 140-240 Cα atoms. f) RMSD of the αG-α4 loop (purple), the distance between K3426.31 and D323Gα (salmon), and the distance between GDP and D295Gα (green) during Gs binding, AHD opening and GDP dissociation; g) and h) comparison between states extracted from before and after AHD opening. Before AHD opening (a), GLP-1R ICL3 interacted with D323Gα and D295 Gα interacted with GDP; after AHD opening (b) and αG-α4 loop reorganization (curved red arrow), αG and D295 Gα moved away from GDP (straight red arrow), destabilizing its binding to Gs.

According to the model of G protein activation, the G protein binding has the effect of allosterically stabilizing the orthosteric agonist in complex with the receptor50 and destabilizing the guanosine-diphosphate (GDP) bound to Gα, triggering its release and exchange with guanosine-triphosphate (GTP)51, upon opening of the G protein alpha-helical domain (AHD). Following this model, PF06882961 and GDP were respectively stabilized and destabilized during the simulated Gs association (Figure S8b,c). The analysis of atomic contacts along the binding path of Gs to GLP-1R highlights a few persistent interactions not observed in the equilibrium MD simulations of GLP-1R:Gs cryo-EM complexes50; for example, we propose the hidden interaction between D3446.33 and R385α5 to be important for Gs coupling (Table S2), which would explain the GLP-1 EC50 reduction upon mutation of position 344 to Ala52.

Extending Replica 2, we further investigated the Gs activation mechanism by supervising the opening of the Gs alpha-helical domain (AHD), which is considered a necessary step to allow GDP release from the Ras-like domain53. We first easily obtained the opening of AHD (Figure 4b) and successively supervised the GDP unbinding in a further three replicas, seeded after the AHD opening. In one of these three mwSuMD simulations, the nucleotide dissociated from Gs (Figure 4d). Video S7 shows the full Gs binding, AHD opening, and GDP release. mwSuMD suggested several structural changes as implicated in GDP dissociation (Figure 4f-h): i) the AHD opening; ii) the conformational change of αG-α4 loop (residues A303Gα-P332Gα), in concert with a loosening of interactions between D323Gα and K3426.31, and iii) the rupture of the hydrogen bond between GDP and D295Gα triggered by the movement of αG away from the GDP binding site (Figure 4f). The involvement of αG through the αG-α4 during the release of GDP from Gs is supported by hydrogen/deuterium exchange experiments54, while the role of D275Gα has been probed with functional assays on the Gi isoform mutant D272GαiA55. Moreover, a similar αG behavior was recently suggested by MD simulations of the Gs binding pathway to β2AR56. We also note that the αG-α4 loop length and aminoacidic composition diverge among G protein isoforms, further suggesting a role in G protein selectivity consistent with the hypothesis that, in different G proteins, distinct domains of the Gα subunit could be responsible for receptor selectivity57. However, a more subtle allosteric communication through internal structural elements like the β2-β3 strands, prompted by α5 tilting, could have weakened GDP phosphate binding as previously suggested by other groups58–60. Interestingly, no significant conformational changes of the β6-α5 loop happened before or during the GDP dissociation, suggesting that its conformational change as captured in the nucleotide-free GLP-1R:Gs complex (Figure S10) occurs after the GDP release as a result of the loss of binding stabilization, rather than being an initiator of the GDP dissociation.

Discussion

Classic MD simulations sample phase space with an efficiency that depends on the energy barrier between neighboring minima. Processes like (un)binding and protein activation require the system to overcome numerous energy barriers, some creating bottlenecks that slow the transition down to the millisecond, or second, time scale. To reduce some of these limits, we have developed, and tested on complex structural events characterizing GPCRs, an energetically unbiased adaptive sampling algorithm, namely multiple walker SuMD, which is based on traditional SuMD, while drawing on parallel multiple replica methods61,62, Our simulations propose that remarkable predictivity can be obtained with distance-driven mwSuMD, as demonstrated by the lowest deviation from the experimental AVP:V2R complex. The dissociation of AVP from V2R was simulated much more rapidly by mwSuMD than by SuMD, suggesting it is an efficient tool for studying the dissociation of ligands from GPCRs. This is due to the more extensive sampling obtainable by seeding multiple parallel short simulations instead of a single simulation for batch.

mwSuMD performed similarly to SuMD for the dynamic docking of AVP to V2R when time windows of 600 ps were employed. Time windows of 100 ps remarkably improved mwSuMD. Usually, dynamic docking is performed to either predict the geometry of complexes or sample the binding path of an already-known intermolecular complex, or both. The RMSD of AVP to the experimental coordinates as the supervised metric produced the best results. Consequently, the RMSD should be the metric of choice to study the binding path of well-known intermolecular complexes. The distance, on the other hand, is necessary when limited structural information about the binding mode is available. In the absence of structural information regarding the final bound state, it is possible to sample numerous binding events employing mwSuMD and evaluate the final bound states rank by applying end-point free energy binding methods like the molecular mechanics energies combined with the Poisson– Boltzmann or generalized Born and surface area continuum solvation (MM/PBSA and MM/GBSA63) models.

We increased the complexity of binding simulations by considering GLP-1R and the non-peptide agonist PF06882961. Using mwSuMD, we obtained a binding of PF06882961in good agreement with the cryo-EM structure, followed by an active-like conformational transition of GLP-1R. The choice of the metrics supervised was driven by the structural data available41 and extensive preparatory MD simulations. However, binding routes are possible from either the bulk solvent or the membrane36,64,65. These results show the power of the mwSuMD method, indicating that future applications could include ligand binding from the membrane, or alternative apo receptor conformations to improve the sampling for more difficult receptors.

mwSuMD enabled us to simulate the Gs binding to the active GLP-1R and the subsequent GDP release. Our results suggest a concerted effect on GDP binding produced by AHD opening, αG-α4 loop rearrangement, and αG shift away from the GDP site. The full rotation and elongation of α5 as in cryo-EM structures would occur after the GDP release, supporting the role of hidden, metastable interactions as the driving force of G protein coupling and selectivity, as per recent work on GLP-1R52.

We stress that a complete understanding of a complex molecular event like G protein coupling requires the collection of numerous G protein binding paths and GDP dissociation events. mwSuMD is designed to yield a mechanistic description of structural events. For this reason, it integrates well with mutagenesis and kinetics experiments. We note that mwSuMD trajectories, since they describe the sequential states along a transition pathway, can represent a precious backbone for further MD sampling aimed at quantifying or predicting the kinetics of the transition. Approaches such as path collective variable metadynamics66, Markov State Models67, or machine learning models68,69 can be informed by mwSuMD. We will address these novel opportunities created by mwSuMD in future work.

In summary, we showcased the applicability domain of mwSuMD to key, but scarcely understood, aspects of GPCR structural biology, pharmacology, and drug design hitherto unaddressed by unbiased simulations. Given the generality and simplicity of its implementation, we anticipate that mwSuMD can be employed to study a wide range of structural phenomena characterizing both membrane and cytosolic proteins. mwSuMD is an ongoing project updated on https://github.com/pipitoludovico/mwSuMD_multiEngine and, in the recent implementation, can exploit ACEMD70, NAMD71, GROMACS72 and OPENMM73 as graphic processing units (GPU)-based MD engines while offering the option to run also on CPUs with GROMACS.

Methods

Multiple walker SuMD (mwSuMD) protocol

The supervised MD (SuMD) is an adaptive sampling method74 based on a tabu-like algorithm for speeding up the simulation of binding events between small molecules (or peptides75,76) and proteins1,19 without the introduction of any energetic bias. Briefly, during SuMD, a series of short unbiased MD simulations are performed, and after each simulation, the distances between the centers of mass (or the geometrical centers) of the ligand and the predicted binding site (collected at regular time intervals) are fitted to a linear function. If the resulting slope is negative (showing progress towards the target), the next simulation step starts from the last set of coordinates and velocities, otherwise, the simulation is restarted by randomly assigning the atomic velocities.

mwSuMD is designed to increase the sampling from a specific configuration by seeding user-decided parallel replicas (walkers) rather than one short simulation as per SuMD. Since one replica for each batch of walkers is always considered productive, mwSuMD gives more control than SuMD on the total wall-clock time used for a simulation. On the flip side, to maximise mwSuMD, it is optimal to assign one walker per GPU, requiring multiple GPUs to be effective. However, modern multi-threaded GPUs can still employ mwSuMD with a smaller cost in GPU performance. In the implementation for ACEMD used in this work, mwSuMD needs as input the initial coordinates of the system as a pdb file, the coordinates, and the atomic velocities of the system from the equilibration stage, the topology file of the system, and all the necessary force field parameters. The user can decide to supervise one (X) or two metrics (X’, X’’) of the simulated system over short simulations seeded in batches, called walkers. In the former case, either the slope of the linear function interpolating the metric values or a score can be adopted to decide whether to continue the mwSuMD simulation. When the user decides to supervise two metrics, a specific score is used. In the present work, distances between centroids, RMSDs, or the number of atomic contacts between two selections were supervised (Table S1). The choice of the metrics is system and problem-dependent, as the RMSD is most useful when the final state is known, while the distance is required when the target state is unknown; details on the scores are given below. The decision to restart or continue mwSuMD after any short simulation is postponed until all the walkers of a batch are collected. The best short simulation is selected and extended by seeding the same number of walkers, with the same duration as the step before.

For each walker, the score for the supervision of a single metric (SMscore) is computed as the square root of the product between the metric value in the last frame (Xlast frame) and the average metric value over the short simulation : If the metric is set to decrease (e.g. binding or dimerization) the walker with the lowest SMscore is continued, otherwise (e.g. unbinding or outwards opening of domains), the walker with the highest score is continued. Using the SMscore rather than the slope should give more weight to the final state of each short simulation, as it is the starting point for the successive batch of simulations. Considering the average of the metric should favor short simulations consistently evolving in the desired direction along the metric.

If both X’ and X’’ are set to increase during the mwSuMD simulations, the score for the supervision of two metrics (DMscore) on each walker is computed as follows: Where X’last frame and X’’last frame are the metrics values in the last frame, while and represent the average value of the two metrics over all the walkers in the batch. Subtracting the value 1 to the metric ratio ensures that if one of the two metrics from the last frame (X’last frame or X’’last frame) is equal to the average then that metric addend is null, and DMscore depends only on the remaining metric. If any of the two metrics is set to decrease, then the corresponding component in Equation 2 is multiplied by - 1 to maintain a positive score. Considering the average value of the two metrics over all the walkers rather than only over the considered walker should be more representative of the system evolution along the defined metric. In other words, the information about the metric is taken from all the walkers to better describe the evolution of the system.

The DMScore is designed to preserve some degree of independence between the two metrics supervised. Indeed, if the variation of one of them slows down and gets close to zero, the other metric is still able to drive the system’s evolution. It should be noted that DMScore works at its best if the two metrics have similar variations over time, as in the case of distance and RMSD (both of which are distance-based). Notably, differently from SuMD, when a walker is extended by seeding a new batch of short simulations and the remaining walkers are stopped, the atomic velocities are not reassigned. This allows the simulations to be as short as a few picoseconds if desired, without introducing artifacts due to the thermostat latency to reach the target temperature (usually up to 10-20 ps when a simulation is restarted, reassigning the velocities of the atoms).

The current implementation of mwSuMD is for Python3 and exploits MDAnalysis77 and MDTRaj78 modules.

Force field, ligands parameters, and general systems preparation

The CHARMM3679,80/CGenFF 3.0.181–83 force field combination was employed in this work. Initial ligand force field, topology and parameter files were obtained from the ParamChem webserver81. Restrained electrostatic potential (RESP)84 partial charges were assigned to all the non-peptidic small molecules but adrenaline and guanosine-5’-diphosphate (GDP) using Gaussian09 (HF/6-31G* level of theory) and AmberTools20.

Six systems were prepared for MD (Table S1). Hydrogen atoms were added using the pdb2pqr85 and propka86 software (considering a simulated pH of 7.0); the protonation of titratable side chains was checked by visual inspection. The resulting receptors were separately inserted in a 1-palmitoyl-2-oleyl-sn-glycerol-3-phosphocholine (POPC) bilayer (previously built by using the VMD Membrane Builder plugin 1.1, Membrane Plugin, Version 1.1. at: http://www.ks.uiuc.edu/Research/vmd/plugins/membrane/), through an insertion method87. Receptor orientation was obtained by superposing the coordinates on the corresponding structure retrieved from the OPM database88. Lipids overlapping the receptor transmembrane helical bundle were removed and TIP3P water molecules89 were added to the simulation box employing the VMD Solvate plugin 1.5 (Solvate Plugin, Version 1.5. at <http://www.ks.uiuc.edu/Research/vmd/plugins/solvate/). Finally, overall charge neutrality was reached by adding Na+/Cl- counter ions up to the final concentration of 0.150 M), using the VMD Autoionize plugin 1.3 (Autoionize Plugin, Version 1.3. at <http://www.ks.uiuc.edu/Research/vmd/plugins/autoionize/).

System equilibration and general MD settings

The MD engine ACEMD370 was employed for both the equilibration and productive simulations. The equilibration was achieved in isothermal-isobaric conditions (NPT) using the Berendsen barostat90 (target pressure 1 atm) and the Langevin thermostat91 (target temperature 300 K) with low damping of 1 ps-1. For the equilibration (integration time step of 2 fs): first, clashes between protein and lipid atoms were reduced through 1500 conjugate-gradient minimization steps, then a positional constraint of 1 kcal mol-1 Å-2 on all heavy atoms was gradually released over different time windows: 2 ns for lipid phosphorus atoms, 60 ns for protein atoms other than alpha carbon atoms, 80 ns for alpha carbon atoms; a further 20 ns of equilibration was performed without any positional constraints.

Productive trajectories (Table S1) were computed with an integration time step of 4 fs in the canonical ensemble (NVT). The target temperature was set at 300 K, using a thermostat damping of 0.1 ps-1; the M-SHAKE algorithm92,93 was employed to constrain the bond lengths involving hydrogen atoms. The cut-off distance for electrostatic interactions was set at 9 Å, with a switching function applied beyond 7.5 Å. Long-range Coulomb interactions were handled using the particle mesh Ewald summation method (PME)94 by setting the mesh spacing to 1.0 Å.

Vasopressin binding simulations

The vasopressin 2 receptor (V2R) in complex with vasopressin (AVP) and the Gs protein95 was retrieved from the Protein Data Bank96 (PDB 7DW9). The Gs was removed from the system and the missing residues on ECL2 (G185-G189) were modeled from scratch using Modeller 9.1997, considering the solution with the lowest DOPE score out of ten conformations produced. AVP was placed away from V2R in the extracellular bulk and the resulting system was prepared for MD simulations and equilibrated as reported above.

During SuMD simulations, the distance between the centroids of AVP residues C1-Q4 (backbone and side chains), anticipated to bind deep into V2R, and the V2R residues lining the peptide binding site Q96, Q174, Q291, and L312 (Cα atoms only) was supervised over time windows of 600 ps or 100 ps (Table S1).mwSuMD simulations considered the same distance, the RMSD of AVP residues C1-Q4 to the experimental bound complex or the combination of the two during time windows of 600 ps (3 walkers) or 100 ps (10 walkers) (Table S1). Slope, SMscore, or DMscore (see Methods section mwSuMD protocol) was used in the different mwSuMD replicas performed (Table S1). Simulations were stopped after 300 ns (time window duration = 600 ps) or 50 ns (time window duration = 100 ps) of total SuMD or mwSuMD simulation time.

Vasopressin unbinding simulations

The V2R:AVP complex was prepared for MD simulations and equilibrated as reported above. During both SuMD and mwSuMD simulations (Table S1), the distance between the centroids of AVP residues C1-Q4 (backbone and side chains) and V2R residues Q96, Q174, Q291, and L312 (Cα atoms only) was supervised over time windows of 100 ps (10 walkers seeded for mwSuMD simulations). Replicas were stopped when the AVP-V2R distance reached 40 Å.

GLP-1R:PF06882961 binding simulations

The inactive glucagon-like peptide receptor (GLP-1R) was retrieved from the Protein Data Bank96 (PDB 6LN2)98. Fab and the intracellular negative allosteric modulator were removed, and missing residues in the stalk (129-134) and ICL2 (256 to 263) were modeled with Modeller 9.19, considering the solutions with the lowest DOPE score out of ten conformations produced. The PF06882961 initial conformation was extracted from the complex with the fully active GLP-1R45 (PDB 7LCJ) and placed away from GLP-1R in the extracellular bulk. The resulting system was prepared for MD simulations and equilibrated as reported above. CGenFF dihedral force field parameters of PF06882961 with the highest penalties (dihedrals NG2R51-CG321-CG3C41-CG3C41 (penalty=143.5) and NG2R51-CG321-CG3C41-OG3C51 (penalty=152.4)) were optimized (Figure S11) employing Gaussian09 (geometric optimization and dihedral scan at HF/6-31g(d) level of theory) and the VMD force field toolkit plugin99.

Four classic MD replicas, for a total of 8 μs, were performed on the inactive receptor (prepared for MD simulations and equilibrated as reported above) to assess the possible binding path to the receptor TMD and therefore decide the initial position of PF06882961 in the extracellular bulk of the simulation box. A visual inspection of the trajectories suggested three major conformational changes that could allow ligand access to the TMD (Figure S12). Transitory openings of the ECD (distance Q47ECD - S310ECL2), TM6-TM7 (distance H3636.52 - F3907.45), and TM1-ECL1 (distance E1381.33 and W214ECL1) were observed. Since the opening of TM1-ECL1 was observed in two replicas out of four, we placed the ligand in a favorable position for crossing that region of GLP-1R.

mwSuMD simulations (Table S1) were performed stepwise to dock the ligand within GLP-1R first and then relax the receptor towards the active state. The PF06882961 binding was obtained by supervising at the same time the distance between the ligand’s heavy atoms centroid and the centroid of GLP-1R TM7 residues L3797.34-F3817.36 (Cα atoms only), which are part of the orthosteric site, and the RMSD of the ECD (residues W33ECD-W120ECD, Cα atoms only) to the active state (PDB 7LCJ) until the former distance reached 4 Å. In the second phase of mwSuMD, the RMSD of the ECD (residues W33ECD-W120ECD, Cα atoms only) and the ECL1 to the active state (PDB 7LCJ, Cα atoms of residues M2042.74-L2243.27) were supervised until the latter reached less than 4 Å. During the third phase, the RMSD of PF06882961, as well as the RMSD of ECL3 (residues A3686.57-T3787.33, Cα atoms), were supervised until the former reached values lower than 3 Å. In the last mwSuMD step, only the RMSD of TM6 (residues I3456.34-F3676.56, Cα atoms) to the active state (PDB 7LCJ) was supervised until less than 5 Å. RMSDs were computed after superimposition on TM2, ECL1, and TM3 residues 170-240 (Cα atoms), which is the GLP-1R less flexible part50.

Membrane-anchored Gs protein:GLP-1R simulations and GDP dissociation

The PDB 6EG8 was processed through Charmm-GUI100 to palmitoylate residue C3Gαi and geranylgeranylate residue C65Gγ. The resulting system was inserted into a 120 x 120 Å POPC membrane and previously built by using the VMD Membrane Builder plugin 1.1, Membrane Plugin, Version 1.1. at: http://www.ks.uiuc.edu/Research/vmd/plugins/membrane. Lipids overlapping the palmitoyl and geranylgeranyl groups were removed and TIP3P water molecules89 were added to the simulation box by means of the VMD Solvate plugin 1.5 (Solvate Plugin, Version 1.5. at <http://www.ks.uiuc.edu/Research/vmd/plugins/solvate/).

Finally, overall charge neutrality was reached by adding Na+/Cl- counter ions up to the final concentration of 0.150 M), using the VMD Autoionize plugin 1.3 (Autoionize Plugin, Version 1.3. at <http://www.ks.uiuc.edu/Research/vmd/plugins/autoionize/). The first stage of equilibration was performed as reported above (Methods section System equilibration and general MD settings) for 120 ns, followed by a second stage in the NVT ensemble for a further 1 μs without any restraints to allow the membrane-anchored heterotrimeric Gs protein to stabilize within the intracellular side of the simulation box. After this two-stage long equilibration, GLP-1R from the final frame of the activation simulation (in complex with PF06882961) was manually inserted into the equilibrated membrane above the Gs protein using the corresponding structure retrieved from the OPM database as a reference, and the system further equilibrated for 120 ns as reported above (Methods section System equilibration and general MD settings). The GLP-1R-Gs system was then subjected to three simulations (Table S1). Each mwSuMD replica was interrupted by 500 ns of classic MD twice, to relax the system during the transition. In the supervised stages, the distance between residues M386-L394 Gαs (all-atoms centroid) of helix 5 (α5) and the GLP-1R intracellular residues R1762.46, R3486.37, S3526.41, and N4057.60 (Cα atoms only) was monitored, seeding three walkers of 200 ps each.

The AHD opening was simulated starting from the GLP-1R:Gs binding mwSuMD replica with the final lowest Gs RMSD, the lowest PF06882961 binding energy and the highest GDP binding energy (Replica 2 in Figure 4e) by supervising the distance between AHD residues G70-R199 Gαs and K300-L394Gαs (all-atoms centroids) during three walkers of 100 ps each. 300 ns of classic MD was performed to relax the system. Finally, the GDP unbinding was supervised as the distance between GDP (all-atoms centroid) and residues E50Gαs, K52Gαs, T55Gαs, K293Gαs, and V367Gαs (Cα atoms only) of Gαs; five walkers were used in a 50 ps long mwSuMD simulations.

MD Analysis

Interatomic distances were computed through MDAnalysis77; root mean square deviations (RMSD) were computed using VMD101 and MDAnalysis77. Interatomic contacts and ligand-protein hydrogen bonds were detected using the GetContacts scripts tool (https://getcontacts.github.io), setting a hydrogen bond donor-acceptor distance of 3.3 Å and an angle value of 120° as geometrical cut-offs. Contacts and hydrogen bond persistency are quantified as the percentage of frames (over all the frames obtained by merging the different replicas) in which protein residues formed contacts or hydrogen bonds with the ligand.

The MMPBSA.py102 script, from the AmberTools20 suite (The Amber Molecular Dynamics Package, at http://ambermd.org/), was used to compute molecular mechanics energies combined with the generalized Born and surface area continuum solvation (MM/GBSA) method or the molecular mechanics Poisson-Boltzmann surface area (MM/PBSA) approach, after transforming the CHARMM psf topology files to an Amber prmtop format using ParmEd (documentation at <http://parmed.github.io/ParmEd/html/index.html).

Supplementary Videos were produced employing VMD and avconv (at https://libav.org/avconv.html). Molecular graphics images were produced using the UCSF Chimera103 (v1.14).

Numbering system

Throughout the manuscript, the Ballesteros-Weinstein residues numbering system for class A104 and the Wootten residues numbering system for class B GPCRs105 are adopted.